---
title: "Group 21 Project"
output:
  pdf_document: default
  word_document: default
  html_document: default
---
```{r setup-data, include=FALSE}
source("helper.r")
source("merging.r")

# install and load packages
install_and_load()

# Load the data frames
customer_frame <- load_data("archive/olist_customers_dataset.csv")

orders_frame <- load_data("archive/olist_orders_dataset.csv")
order_items_frame <- load_data("archive/olist_order_items_dataset.csv")
payment_frame <- load_data("archive/olist_order_payments_dataset.csv")

product_frame <- load_data("archive/olist_products_dataset.csv")

translation_frame <- load_data("archive/product_category_name_translation.csv")

reviews_frame <- load_data("archive/olist_order_reviews_dataset.csv")
sellers_frame <- load_data("archive/olist_sellers_dataset.csv")

orders_enriched <- merge_olist_data(
  orders_frame,
  translation_frame,
  order_items_frame,
  payment_frame,
  customer_frame,
  sellers_frame,
  product_frame
)

late_rate <- round(mean(orders_enriched$is_late == "Late") * 100, 1)
```

# Introduction of the Business Problem

## The Business Context
Olist operates as a multi-seller marketplace. Like many e-commerce platforms, Olist suffers from delayed deliveries (Across the dataset, `r late_rate`% of delivered orders arrived after the estimated delivery date.), which inevitably leads to poor customer reviews, platform churn, and high customer support costs. Because Olist does not own the inventory or directly control the dispatch process, the root cause of these delays often stems from individual seller behavior (e.g., poor logistics management, slow processing times). 

## The Problem
Currently, Olist's approach to delayed deliveries is purely reactive—dealing with angry customers *after* the estimated delivery date has already been missed. To reduce support costs and protect the platform's brand reputation, Olist needs a proactive mechanism to identify at-risk orders before they fail, and a strategic framework to manage underperforming sellers.

```{r proof-of-concept, echo=FALSE}

poc_frame <- merge(
  orders_enriched, 
  reviews_frame[, c("order_id", "review_score")], 
  by = "order_id", 
  all.x = TRUE
)

# Remove any rows where review_score is NA (if a customer didn't leave a review)
poc_frame <- poc_frame[!is.na(poc_frame$review_score), ]

# 5. Aggregate to find the average review score for Late vs. On-Time deliveries
review_impact <- aggregate(
  review_score ~ is_late, 
  data = poc_frame, 
  FUN = function(x) c(mean = mean(x), count = length(x))
)

# Flatten the aggregate output for clean printing
review_impact <- do.call(data.frame, review_impact)
colnames(review_impact) <- c("Delivery_Status", "Average_Review_Score", "Total_Orders")


# 6. Visualize the Proof
barplot(
  review_impact$Average_Review_Score, 
  names.arg = review_impact$Delivery_Status, 
  col = c("firebrick", "steelblue"),
  main = "Average Review Score: Late vs. On-Time Deliveries",
  ylab = "Average Star Rating (1-5)",
  ylim = c(0, 5)
)

```

## The Analytics Task
We aim to adopt a two-stage analytical approach to transition Olist from reactive to proactive logistics management:

1. **Unsupervised Learning (Strategic Seller Segmentation):** First, we will engineer features to capture historical logistics performance and pricing behavior at the *seller level* (e.g., average delay rate, average freight cost, dispatch speed). We will use **K-Means Clustering** to segment our sellers into distinct performance tiers (e.g., "High-Reliability," "Slow & Cheap," "High-Risk").
2. **Supervised Learning (Tactical Order Forecasting):** Second, we will map these seller segments back to the individual order level. We will use these segments as a key predictor in a classification model (such as a Random Forest or Logistic Regression) to forecast whether a *future* individual order is likely to be delivered late.

**Business Impact:** This pipeline enables a dual-pronged business strategy. Strategically, Olist can use the seller clusters to penalize or coach "High-Risk" sellers (addressing the root cause). Tactically, the predictive model allows Olist's customer service team to identify high-risk orders the moment they are placed, enabling proactive communication with the customer or the deployment of expedited shipping interventions (addressing the symptoms).

## Data dictionary
```{r data-table, echo=FALSE}
# Create a data dictionary of the merged dataset
data_dictionary <- data.frame(
  variable = names(orders_enriched),
  data_role = c(
    "identifier", "identifier", "identifier", "identifier",
    "temporal",
    "predictor", "predictor", "predictor", "predictor",
    "predictor", "predictor", "predictor",
    "context", "context", "context",
    "identifier", "context",
    "temporal", "temporal", "temporal", "temporal", "temporal",
    "predictor", "predictor",
    "identifier", "context", "context", "context",
    "outcome", "outcome"
  ),
  description = c(
    "Unique identifier for each order",
    "Unique identifier for the seller",
    "Unique identifier for the product",
    "Sequence number of item within the order",
    "Seller's shipping deadline for the item",
    "Listed price of the product",
    "Freight/shipping cost for the item",
    "Weight of the product in grams",
    "Length of the product in centimetres",
    "Height of the product in centimetres",
    "Width of the product in centimetres",
    "Product category name (English)",
    "Seller's zip code prefix",
    "Seller's city",
    "Seller's state",
    "Unique identifier for the customer (order-level)",
    "Order status at time of data extraction",
    "Timestamp when the order was placed",
    "Timestamp when payment was approved",
    "Date the order was handed to the carrier",
    "Date the order was delivered to the customer",
    "Estimated delivery date shown to customer",
    "Total payment value for the order (aggregated)",
    "Maximum instalment plan length for the order",
    "Anonymised unique customer identifier",
    "Customer's zip code prefix",
    "Customer's city",
    "Customer's state",
    "Target variable: whether the order was delivered late",
    "Signed delivery delay in days (negative = early)"
  )
)
knitr::kable(
  data_dictionary,
  format = "markdown",
  col.names = c("Variable", "Data Role", "Description")
)
```
# Data Preparation

## Dataset Overview and Selection

The Olist dataset comprises eight relational tables linked by shared keys. We
load all eight and make deliberate inclusion decisions based on relevance to
the business problem. The table below summarises these decisions.

| Dataset | Included | Rationale |
|---------|----------|-----------|
| olist_orders | Yes | Analytical spine — contains all delivery timestamps and order status |
| olist_order_items | Yes | Links sellers to orders; provides price, freight, and product keys |
| olist_sellers | Yes | Seller geography — state and city for Stage 1 clustering |
| olist_customers | Yes | Customer state — used as a predictor in Stage 2 classification |
| olist_products | Yes | Physical dimensions (weight, size) — predictors of delivery complexity |
| olist_order_payments | Yes | Payment value and instalment behaviour — order-level predictors |
| olist_geolocation | No | Ambiguous deduplication required (multiple coordinates per zip code prefix); flagged as a direction for future improvement |
| olist_order_reviews | No | Outcome of delivery performance, not a predictor — inclusion would constitute data leakage. Used only for the proof-of-concept visualisation above. |

## Merge Strategy

The full merge sequence is:

1. `orders` + `payments_aggregated` → via `order_id`
2. + `customers` → via `customer_id`
3. `order_items` + `products_translated` → via `product_id`
4. + `sellers` → via `seller_id`
5. Final join: orders spine + enriched order items → via `order_id`

## Column Removal

## Target Variable Construction
```{r target-variable-summary, echo=FALSE}
cat("On time:", sum(orders_enriched$is_late == "On_Time"), "\n")
cat("Late:   ", sum(orders_enriched$is_late == "Late"), "\n")
cat("Late rate:", round(mean(orders_enriched$is_late == "Late") * 100, 1), "%\n")
cat("Total labelled orders:", nrow(orders_enriched), "\n")
```

# Exploration

## Missing values
```{r missing-values}
# Summary of missing values
missing_summary <- data.frame(
  variable = names(orders_enriched),
  n_missing = sapply(orders_enriched, function(x) sum(is.na(x))),
  pct_missing = sapply(orders_enriched, function(x) round(mean(is.na(x
)) * 100, 2))
)
missing_summary <- missing_summary[order(-missing_summary$pct_missing), ]
print(missing_summary)
```
Since these correspond to products with no physical measurements recorded in the source they will be imputed with the median value for each dimension.
```{r impute-dimensions}
dim_cols <- c("product_weight_g", "product_length_cm",
              "product_height_cm", "product_width_cm")

for (col in dim_cols) {
  orders_enriched[[col]][is.na(orders_enriched[[col]])] <-
    median(orders_enriched[[col]], na.rm = TRUE)
}

cat("Remaining NAs after imputation:", sum(is.na(orders_enriched)), "\n")
```

## Distributions and Skewness

### Numerical Variables
```{r outliers-skewness-distributions}

# check skewness
num_vars <- orders_enriched[, sapply(orders_enriched, is.numeric)]

skewness_summary <- data.frame(
  variable = names(num_vars),
  mean     = sapply(num_vars, mean, na.rm = TRUE),
  median   = sapply(num_vars, median, na.rm = TRUE),
  sd       = sapply(num_vars, sd, na.rm = TRUE),
  skewness = sapply(num_vars, skewness, na.rm = TRUE)
)
print(skewness_summary)

# DISTRIBUTIONS: NUMERICAL VARIABLES
# Faceted histograms for all numeric variables at once
num_long <- tidyr::pivot_longer(num_vars, everything(), 
                                 names_to = "variable", values_to = "value")

ggplot(num_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distribution of Numerical Variables", x = NULL, y = "Count") +
  theme_minimal()

# LOG TRANSFORMATION CHECK for skewed vars
# Visual before/after for price, payment_value_total, freight_value, and product_weight_g 
par(mfrow = c(2, 2))
hist(orders_enriched$price, breaks = 50, main = "Price (raw)", xlab = "")
hist(log1p(orders_enriched$price), breaks = 50, main = "Price (log)", xlab = "")
hist(orders_enriched$payment_value_total, breaks = 50, main = "Payment (raw)", xlab = "")
hist(log1p(orders_enriched$payment_value_total), breaks = 50, main = "Payment (log)", xlab = "")
hist(orders_enriched$freight_value, breaks = 50, main = "Freight (raw)", xlab = "")
hist(log1p(orders_enriched$freight_value), breaks = 50, main = "Freight (log)", xlab = "")
hist(orders_enriched$product_weight_g, breaks = 50, main = "Weight (raw)", xlab = "")
hist(log1p(orders_enriched$product_weight_g), breaks = 50, main = "Weight (log)", xlab = "")
par(mfrow = c(1, 1))

```
### Categorical variables
```{r categorical-distributions, fig.height=5}
par(mfrow = c(1, 3))

state_counts <- sort(table(orders_enriched$customer_state), decreasing = TRUE)
barplot(state_counts, las = 2, main = "Orders by Customer State",
        col = "steelblue", cex.names = 0.7, ylab = "Count")

seller_state_counts <- sort(table(orders_enriched$seller_state), decreasing = TRUE)
barplot(seller_state_counts, las = 2, main = "Sellers by State",
        col = "coral", cex.names = 0.7, ylab = "Count")

cat_counts <- sort(table(orders_enriched$product_category_name_english),
                   decreasing = TRUE)
barplot(head(cat_counts, 20), las = 2, main = "Top 20 Product Categories",
        col = "steelblue", cex.names = 0.55, ylab = "Count")

par(mfrow = c(1, 1))
```

Both customer and seller activity are heavily concentrated in São Paulo (SP),
which accounts for approximately 40% of orders and the plurality of sellers.
This geographic concentration is relevant for our clustering stage — SP-based
sellers may exhibit systematically different logistics profiles to sellers in
more remote states. Product categories show a similarly long-tail distribution,
with bed/bath/table and health & beauty dominating by volume. We retain the
raw category variable as a predictor in the classification model without
further consolidation, as Random Forest can handle high-cardinality categoricals
natively.


## Data transformation


## Numerical variable transformations
```{r numerical-transformations}
# check skewness before transformation
skewness(orders_enriched$price, na.rm = TRUE)
skewness(orders_enriched$payment_value_total, na.rm = TRUE)
skewness(orders_enriched$freight_value, na.rm = TRUE)
skewness(orders_enriched$product_weight_g, na.rm = TRUE)
# Log transformation for skewed variables
orders_enriched$price_log <- log1p(orders_enriched$price)
orders_enriched$payment_value_total_log <- log1p(orders_enriched$payment_value_total)
orders_enriched$freight_value_log <- log1p(orders_enriched$freight_value)
orders_enriched$product_weight_g_log <- log1p(orders_enriched$product_weight_g)
# Check skewness after transformation
skewness(orders_enriched$price_log, na.rm = TRUE)
skewness(orders_enriched$payment_value_total_log, na.rm = TRUE)
skewness(orders_enriched$freight_value_log, na.rm = TRUE)
skewness(orders_enriched$product_weight_g_log, na.rm = TRUE)
```

# Feature Engineering

## Seller-Level Feature Engineering

The first stage of our analytical pipeline requires building a performance 
profile for each seller. Rather than analysing individual orders in isolation, 
we aggregate historical order data to the seller level, capturing each seller's 
logistics behaviour across all their past orders. These features will serve as 
the input to our K-Means clustering model.

We engineer four features that capture distinct dimensions of seller performance:

- **Late rate:** the proportion of a seller's orders delivered after the 
  estimated date — our primary measure of reliability
- **Avg dispatch days:** the average number of days between order purchase and 
  carrier handoff — captures how quickly a seller processes and ships orders
- **Avg freight cost:** the average shipping cost charged across a seller's 
  orders — a proxy for logistics complexity and geographic reach
- **Order volume:** the total number of fulfilled orders — distinguishes 
  established sellers from low-activity ones whose metrics may be unreliable

```{r seller-feature-engineering}
# Build a seller-level performance profile from order-level data

# binary late flag at order level (numeric for aggregation)
orders_enriched$is_late_numeric <- ifelse(
  orders_enriched$is_late == "Late", 1, 0
)

# (days between purchase and carrier handoff)
orders_enriched$dispatch_days <- as.numeric(difftime(
  orders_enriched$order_delivered_carrier_date,
  orders_enriched$order_purchase_timestamp,
  units = "days"
))

# Aggregate to seller level
seller_profiles <- data.frame(
  seller_id = unique(orders_enriched$seller_id)
)

# Late delivery rate per seller
seller_profiles$late_rate <- tapply(
  orders_enriched$is_late_numeric,
  orders_enriched$seller_id,
  mean, na.rm = TRUE
)[seller_profiles$seller_id]

# Average dispatch speed per seller
seller_profiles$avg_dispatch_days <- tapply(
  orders_enriched$dispatch_days,
  orders_enriched$seller_id,
  mean, na.rm = TRUE
)[seller_profiles$seller_id]

# Average freight cost per seller
seller_profiles$avg_freight_log <- tapply(
  orders_enriched$freight_value_log,
  orders_enriched$seller_id,
  mean, na.rm = TRUE
)[seller_profiles$seller_id]

# Order volume per seller
seller_profiles$order_volume <- tapply(
  orders_enriched$is_late_numeric,
  orders_enriched$seller_id,
  length
)[seller_profiles$seller_id]

# Remove sellers with fewer than 10 orders
# Sellers with very few orders produce statistically unreliable rate estimates
# (e.g. 1 late delivery out of 2 orders = 50% late rate, which is misleading)
cat("Sellers before volume filter:", nrow(seller_profiles), "\n")
seller_profiles <- seller_profiles[seller_profiles$order_volume >= 10, ]
cat("Sellers after volume filter:", nrow(seller_profiles), "\n")
cat("Orders covered by remaining sellers:", 
    sum(orders_enriched$seller_id %in% seller_profiles$seller_id), "\n")

# Check for missing values in seller profiles
cat("\nMissing values in seller profiles:\n")
print(colSums(is.na(seller_profiles)))

# Inspect the distributions
summary(seller_profiles[, c("late_rate", "avg_dispatch_days", 
                              "avg_freight_log", "order_volume")])
```

The distributions of these seller features reveal important patterns before 
we proceed to clustering:
```{r seller-profile-distributions}
par(mfrow = c(2, 2))

hist(seller_profiles$late_rate, breaks = 30,
     main = "Late Delivery Rate", 
     xlab = "Proportion of Late Orders",
     col = "steelblue")

hist(seller_profiles$avg_dispatch_days, breaks = 30,
     main = "Average Dispatch Speed",
     xlab = "Days from Purchase to Carrier",
     col = "steelblue")

hist(seller_profiles$avg_freight_log, breaks = 30,
     main = "Average Freight Cost",
     xlab = "R$ Freight Value",
     col = "steelblue")

hist(seller_profiles$order_volume, breaks = 30,
     main = "Order Volume",
     xlab = "Number of Orders",
     col = "steelblue")

par(mfrow = c(1, 1))
```

## Scaling for Clustering

K-Means is a distance-based algorithm and is therefore sensitive to the scale 
of input features. A seller with an average freight cost of R$50 would 
dominate distance calculations over a late rate expressed as a proportion 
between 0 and 1, simply due to the difference in units rather than any 
meaningful difference in importance. We therefore standardise all four 
features to zero mean and unit variance before clustering.
```{r seller-profile-scaling}
# Remove NAs before scaling
seller_profiles_clean <- na.omit(seller_profiles)
cat("Sellers lost to NA removal:", 
    nrow(seller_profiles) - nrow(seller_profiles_clean), "\n")

# Scale features (exclude seller_id)
seller_scaled <- scale(
  seller_profiles_clean[, c("late_rate", "avg_dispatch_days",
                              "avg_freight_log", "order_volume")]
)

# Confirm zero mean and unit variance
round(apply(seller_scaled, 2, mean), 6)
round(apply(seller_scaled, 2, sd), 3)
```

## PCA Feasibility Assessment

Before clustering, we assess whether dimensionality reduction via PCA is 
appropriate. With only four features, PCA is unlikely to dramatically simplify 
the input space — however it serves two purposes here: it reveals how much 
variance each feature contributes, and it checks whether any features are 
redundant (highly correlated), which would justify exclusion before clustering.
```{r pca-assessment}
pca_result <- prcomp(seller_scaled, center = FALSE, scale. = FALSE)

# Variance explained by each component
pca_summary <- summary(pca_result)
print(pca_summary)

# Scree plot
plot(pca_summary$importance[2, ], 
     type = "b", pch = 19,
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot — Seller Profile PCA")

# Loadings: which features drive each component?
print(pca_result$rotation)

# Correlation matrix to check for redundancy
cor_matrix <- cor(seller_profiles_clean[, c("late_rate", "avg_dispatch_days",
                                              "avg_freight_log", "order_volume")])
print(round(cor_matrix, 3))
```

If the first two components explain more than 80% of variance, we use the PCA 
scores as clustering input. If not — meaning all four features contribute 
independently — we proceed with the full scaled feature set, as discarding 
features would lose meaningful signal. We reflect on this decision after 
inspecting the scree plot output.

## K-Means Seller Clustering

With seller profiles scaled and PCA assessed, we now cluster sellers into 
performance tiers using K-Means. We determine the optimal number of clusters 
using the elbow method on total within-cluster sum of squares.
```{r seller-clustering}
set.seed(42)

# Elbow method — test k from 1 to 10
wss <- sapply(1:10, function(k) {
  kmeans(seller_scaled, centers = k, nstart = 25)$tot.withinss
})

plot(1:10, wss, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Total Within-Cluster SS",
     main = "Elbow Method for Optimal k")

# Fit final model — adjust k based on elbow plot
set.seed(42)
k_optimal <- 4  # update after inspecting elbow plot
km_sellers <- kmeans(seller_scaled, centers = k_optimal, nstart = 25)

# Assign cluster labels back to seller profiles
seller_profiles_clean$cluster <- km_sellers$cluster

# Inspect cluster means to interpret what each cluster represents
cluster_means <- aggregate(
  cbind(late_rate, avg_dispatch_days, avg_freight_log, order_volume) ~ cluster,
  data = seller_profiles_clean,
  FUN = mean
)
print(round(cluster_means, 3))

# Cluster sizes
table(seller_profiles_clean$cluster)
```
```{r seller-cluster-interpretation}
# Visualise clusters using factoextra template
clusters <- km_sellers
cluster_plot_data <- seller_profiles_clean[, !names(seller_profiles_clean) %in% c("seller_id", "cluster"), drop = FALSE]

factoextra::fviz_cluster(
  object = clusters,
  data = cluster_plot_data,
  geom = "point",
  ellipse.type = "convex",
  ggtheme = theme_bw()
)

# Assign interpretable names based on cluster_means output

seller_profiles_clean$cluster_label <- dplyr::case_when(
  # bc of the high freight cost
  seller_profiles_clean$cluster == 1 ~ "Premium_Freight",
  # bc of the high late rate and slow dispatch
  seller_profiles_clean$cluster == 2 ~ "High_Risk",
  # bc of the low late rate and fast dispatch, even if freight is not the cheapest
  seller_profiles_clean$cluster == 3 ~ "High_Reliability",
  # bc of the high order volume but average late rate and dispatch speed
  seller_profiles_clean$cluster == 4 ~ "High_Volume",
  TRUE ~ "Unknown"
)

cat("\nSeller segment distribution:\n")
print(table(seller_profiles_clean$cluster_label))
```

## Mapping Seller Segments Back to Orders

The seller cluster labels are now joined back to the main orders dataset, 
making the seller's performance tier available as a feature in the 
classification model. This is the key link between Stage 1 (unsupervised 
seller segmentation) and Stage 2 (supervised order-level prediction).
```{r map-clusters-to-orders}
# Join cluster labels onto orders
orders_enriched <- merge(
  orders_enriched,
  seller_profiles_clean[, c("seller_id", "cluster_label")],
  by = "seller_id",
  all.x = TRUE
)

# Orders from sellers excluded by the volume filter will have NA cluster
cat("Orders with cluster assigned:", 
    sum(!is.na(orders_enriched$cluster_label)), "\n")
cat("Orders without cluster (low-volume sellers):", 
    sum(is.na(orders_enriched$cluster_label)), "\n")

# Assign low-volume sellers their own segment rather than dropping them
orders_enriched$cluster_label[is.na(orders_enriched$cluster_label)] <- "Low_Volume"

# Check late rate by cluster — does segmentation separate risk levels?
aggregate(is_late_numeric ~ cluster_label, 
          data = orders_enriched, 
          FUN = mean)
```

The validation confirms the clustering has successfully separated seller risk levels. The High_Risk segment shows a late delivery rate of 16.2% — more than three times the 5.2% observed in the High_Reliability segment. This suggests that the seller-level features we engineered are meaningful predictors of delivery performance, and that our clustering has effectively captured this signal in a way that can be leveraged for order-level forecasting.

# Modeling

## Model 1 — Random Forest

## Model 2 Logistic Regression

## Comparison

# Conclusion
