---
title: "Group 21 Project"
output:
  pdf_document: default
  word_document: default
  html_document: default
---
```{r setup-data, include=FALSE}
source("helper.r")
source("merging.r")

# install and load packages
install_and_load()

# Load the data frames
customer_frame <- load_data("archive/olist_customers_dataset.csv")

orders_frame <- load_data("archive/olist_orders_dataset.csv")
order_items_frame <- load_data("archive/olist_order_items_dataset.csv")
payment_frame <- load_data("archive/olist_order_payments_dataset.csv")

product_frame <- load_data("archive/olist_products_dataset.csv")

translation_frame <- load_data("archive/product_category_name_translation.csv")

reviews_frame <- load_data("archive/olist_order_reviews_dataset.csv")
sellers_frame <- load_data("archive/olist_sellers_dataset.csv")

orders_enriched <- merge_olist_data(
  orders_frame,
  translation_frame,
  order_items_frame,
  payment_frame,
  customer_frame,
  sellers_frame,
  product_frame
)
```

# Introduction of the Business Problem

## The Business Context
Olist operates as a multi-seller marketplace. Like many e-commerce platforms, Olist suffers from delayed deliveries, which inevitably leads to poor customer reviews, platform churn, and high customer support costs. Because Olist does not own the inventory or directly control the dispatch process, the root cause of these delays often stems from individual seller behavior (e.g., poor logistics management, slow processing times). 

## The Problem
Currently, Olist's approach to delayed deliveries is purely reactive—dealing with angry customers *after* the estimated delivery date has already been missed. To reduce support costs and protect the platform's brand reputation, Olist needs a proactive mechanism to identify at-risk orders before they fail, and a strategic framework to manage underperforming sellers.

```{r proof-of-concept, echo=FALSE}

poc_frame <- merge(
  orders_enriched, 
  reviews_frame[, c("order_id", "review_score")], 
  by = "order_id", 
  all.x = TRUE
)

# Remove any rows where review_score is NA (if a customer didn't leave a review)
poc_frame <- poc_frame[!is.na(poc_frame$review_score), ]

# 5. Aggregate to find the average review score for Late vs. On-Time deliveries
review_impact <- aggregate(
  review_score ~ is_late, 
  data = poc_frame, 
  FUN = function(x) c(mean = mean(x), count = length(x))
)

# Flatten the aggregate output for clean printing
review_impact <- do.call(data.frame, review_impact)
colnames(review_impact) <- c("Delivery_Status", "Average_Review_Score", "Total_Orders")


# 6. Visualize the Proof
barplot(
  review_impact$Average_Review_Score, 
  names.arg = review_impact$Delivery_Status, 
  col = c("firebrick", "steelblue"),
  main = "Average Review Score: Late vs. On-Time Deliveries",
  ylab = "Average Star Rating (1-5)",
  ylim = c(0, 5)
)
```

## The Analytics Task
We aim to adopt a two-stage analytical approach to transition Olist from reactive to proactive logistics management:

1. **Unsupervised Learning (Strategic Seller Segmentation):** First, we will engineer features to capture historical logistics performance and pricing behavior at the *seller level* (e.g., average delay rate, average freight cost, dispatch speed). We will use **K-Means Clustering** to segment our sellers into distinct performance tiers (e.g., "High-Reliability," "Slow & Cheap," "High-Risk").
2. **Supervised Learning (Tactical Order Forecasting):** Second, we will map these seller segments back to the individual order level. We will use these segments as a key predictor in a classification model (such as a Random Forest or Logistic Regression) to forecast whether a *future* individual order is likely to be delivered late.

**Business Impact:** This pipeline enables a dual-pronged business strategy. Strategically, Olist can use the seller clusters to penalize or coach "High-Risk" sellers (addressing the root cause). Tactically, the predictive model allows Olist's customer service team to identify high-risk orders the moment they are placed, enabling proactive communication with the customer or the deployment of expedited shipping interventions (addressing the symptoms).

## Data dictionary
```{r data-table, echo=FALSE}
# Create a data dictionary of the merged dataset
data_dictionary <- data.frame(
  variable = names(orders_enriched),
  data_role = c(
    "identifier",
    "identifier",
    "identifier",
    "identifier",
    "temporal",
    "predictor",
    "predictor",
    "predictor",
    "predictor",
    "predictor",
    "predictor",
    "predictor",
    "context",
    "context",
    "context",
    "identifier",
    "context",
    "temporal",
    "temporal",
    "temporal",
    "temporal",
    "temporal",
    "predictor",
    "predictor",
    "identifier",
    "context",
    "context",
    "context",
    "outcome",
    "outcome"
  ),
  description = c(
    "Unique identifier for each order",
    "Unique identifier for the seller",
    "Unique identifier for the product",
    "Unique identifier for each item in the order",
    "Shipping deadline date",
    "Price of the product in the order item",
    "Freight/shipping cost for the order item",
    "Weight of the product in grams",
    "Length of the product in centimeters",
    "Height of the product in centimeters",
    "Width of the product in centimeters",
    "Product category name in English",
    "Seller's zip code prefix",
    "Seller's city",
    "Seller's state",
    "Unique identifier for each customer",
    "Status of the order",
    "Date and time when the order was placed",
    "Date and time when the order was approved",
    "Date when the order was delivered by carrier",
    "Date when the order was delivered to the customer",
    "Estimated delivery date",
    "Total payment value for the order",
    "Maximum number of installments for payment",
    "Unique identifier for the customer",
    "Customer's zip code prefix",
    "Customer's city",
    "Customer's state",
    "Binary flag indicating if the order was delivered late",
    "Number of days the delivery was delayed"
  )
)
knitr::kable(data_dictionary, format = "markdown", col.names = c("Variable", "Data Role", "Description"))
```

# Data Preparation


# excluding frames
Add text later 

# Exploration

## Missing values
```{r missing-values}
# Summary of missing values
missing_summary <- data.frame(
  variable = names(orders_enriched),
  n_missing = sapply(orders_enriched, function(x) sum(is.na(x))),
  pct_missing = sapply(orders_enriched, function(x) round(mean(is.na(x
)) * 100, 2))
)
missing_summary <- missing_summary[order(-missing_summary$pct_missing), ]
print(missing_summary)
```

## Outliers, skewness, and distributions

### numerical variables
```{r outliers-skewness-distributions}

# Check balance — important for classification
cat("On time:", sum(orders_frame$is_late == 0), "\n")
cat("Late:   ", sum(orders_frame$is_late == 1), "\n")
cat("Late rate:", round(mean(orders_frame$is_late == 1) * 100, 1), "%\n")

# check skewness
num_vars <- orders_enriched[, sapply(orders_enriched, is.numeric)]

skewness_summary <- data.frame(
  variable = names(num_vars),
  mean     = sapply(num_vars, mean, na.rm = TRUE),
  median   = sapply(num_vars, median, na.rm = TRUE),
  sd       = sapply(num_vars, sd, na.rm = TRUE),
  skewness = sapply(num_vars, skewness, na.rm = TRUE)
)
print(skewness_summary)

# DISTRIBUTIONS: NUMERICAL VARIABLES
# Faceted histograms for all numeric variables at once
num_long <- tidyr::pivot_longer(num_vars, everything(), 
                                 names_to = "variable", values_to = "value")

ggplot(num_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distribution of Numerical Variables", x = NULL, y = "Count") +
  theme_minimal()

# LOG TRANSFORMATION CHECK for skewed vars
# Visual before/after for price and payment_value_total
par(mfrow = c(2, 2))
hist(orders_enriched$price, breaks = 50, main = "Price (raw)", xlab = "")
hist(log1p(orders_enriched$price), breaks = 50, main = "Price (log)", xlab = "")
#hist(orders_enriched$payment_value_total, breaks = 50, main = "Payment (raw)", xlab = "")
#hist(log1p(orders_enriched$payment_value_total), breaks = 50, main = "Payment (log)", xlab = "")
par(mfrow = c(1, 1))

#num_vars <- orders_enriched[, c("price", "payment_value_total", "payment_installments_total")]

#boxplot(
 # list(
  #  Price = orders_enriched$price,
    #Payment_Value = orders_enriched$payment_value_total,
    #Payment_Installments = orders_enriched$payment_installments_total
  #),
  #main = "Outliers Across Key Numeric Variables",
  #ylab = "Value"
#
#cor(num_vars, use = "complete.obs")
```
Price and payment value are heavily right-skewed (skewness > 3), suggesting a log transformation may be appropriate before feeding these into distance-sensitive models such as k-means or PCA. The variable customer_state has 27 unique values with SP dominating (~40% of orders), which may warrant grouping by region rather than state in subsequent modelling to avoid sparsity. No variables exceed 1% missingness, so imputation is not strictly necessary.

### Categorical variables

## Data transformation and outlier handling
Based on the skewness and outliers observed in the numerical variables, we will apply a log transformation to the price and payment_value_total variables to reduce skewness and mitigate the influence of extreme values. This will help improve the performance of distance-based models and ensure that the data is more normally distributed for modeling. We will also consider grouping the customer_state variable into broader regions to reduce sparsity and improve model performance, especially for models that may struggle with high-cardinality categorical variables.

## Numerical variable transformations
```{r numerical-transformations}
# Log transformation for skewed variables
orders_enriched$price_log <- log1p(orders_enriched$price)
orders_enriched$payment_value_total_log <- log1p(orders_enriched$payment_value_total)
# Check skewness after transformation
skewness(orders_enriched$price_log, na.rm = TRUE)
skewness(orders_enriched$payment_value_total_log, na.rm = TRUE)
```

## Proving the Business Problem: The Impact of Late Deliveries
Before building predictive models, we must empirically validate our core business assumption: do delayed deliveries actually drive bad reviews? 

First, we will merge the review data, convert our timestamp columns to date objects, and calculate the exact delay for each order.