---
title: "Group 21 Project"
output:
  pdf_document: default
  word_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup-data, include=FALSE}
source("helper.r")
source("merging.r")

# install and load packages
install_and_load()

# Load the data frames
customer_frame <- load_data("archive/olist_customers_dataset.csv")

orders_frame <- load_data("archive/olist_orders_dataset.csv")
order_items_frame <- load_data("archive/olist_order_items_dataset.csv")
payment_frame <- load_data("archive/olist_order_payments_dataset.csv")

product_frame <- load_data("archive/olist_products_dataset.csv")

translation_frame <- load_data("archive/product_category_name_translation.csv")

reviews_frame <- load_data("archive/olist_order_reviews_dataset.csv")
sellers_frame <- load_data("archive/olist_sellers_dataset.csv")

orders_enriched <- merge_olist_data(
  orders_frame,
  translation_frame,
  order_items_frame,
  payment_frame,
  customer_frame,
  sellers_frame,
  product_frame
)

late_rate <- round(mean(orders_enriched$is_late == "Late") * 100, 1)
```

# 1. Introduction

Data analytics has become an increasingly integral part of a company's
decision making process. Utilizing data to uncover patterns, identify
opportunities, and monitor business performance are practices synonymous
with modern corporations. This document shall cover the various
statistical and modeling topics covered in preparation for the Business
Analytics course, and shall relate these to real-world applications.

## Business Context

The modern world has become progressively digitized, with this the share
of online/eCommerce marketplaces have garnered an ever increasing
percentage of the global retail sales market share. As of the writing of
this document the percentage lies around 22.2% (Buck, 2026). With growth
consistently higher than retail, one of the major countries of growth is
Brazil. Attaining a growth rate of 15.4% it was the 5th largest growth
country in 2023 (Oberlo, 2024).

One of the major eCommerce platforms in Brazil is Olist. Like many
eCommerce platforms, Olist suffers from delayed deliveries (Across the
dataset, `r late_rate`% of delivered orders arrived after the estimated
delivery date.), which inevitably leads to poor customer reviews,
platform churn, and high customer support costs. Because Olist does not
own the inventory or directly control the dispatch process, the root
cause of these delays often stems from individual seller behavior (e.g.,
poor logistics management, slow processing times).

## Problem Statement

Currently, Olist's approach to delayed deliveries is purely reactive,
dealing with angry customers only *after* the estimated delivery date
has already been missed. This makes the process inherently suboptimal,
with delays leaving customers unsattisfied. While eCommerce platforms
are largely platforms for outside sellers to place, market, and sell
their products, in the case of an unfavorable experience it is still
Olist which suffers reputational damage.

```{r proof-of-concept, echo=FALSE}

poc_frame <- merge(
  orders_enriched, 
  reviews_frame[, c("order_id", "review_score")], 
  by = "order_id", 
  all.x = TRUE
)

# Remove any rows where review_score is NA (if a customer didn't leave a review)
poc_frame <- poc_frame[!is.na(poc_frame$review_score), ]

# 5. Aggregate to find the average review score for Late vs. On-Time deliveries
review_impact <- aggregate(
  review_score ~ is_late, 
  data = poc_frame, 
  FUN = function(x) c(mean = mean(x), count = length(x))
)

# Flatten the aggregate output for clean printing
review_impact <- do.call(data.frame, review_impact)
colnames(review_impact) <- c("Delivery_Status", "Average_Review_Score", "Total_Orders")


# 6. Visualize the Proof
barplot(
  review_impact$Average_Review_Score, 
  names.arg = review_impact$Delivery_Status, 
  col = c("firebrick", "steelblue"),
  main = "Average Review Score: Late vs. On-Time Deliveries",
  ylab = "Average Star Rating (1-5)",
  ylim = c(0, 5)
)

```

## Project Goal

With the context laid out Problem Statement, the goal of this project is
therefore to reduce support costs and protect the platform's brand
reputation. More specifically this shall be done by identifying where
Olist needs to implement a proactive mechanism to identify at-risk
orders before they fail, as well as a strategic framework to manage
underperforming sellers.

# 2. Data Analysis Giannis/Jonas

Having covered the problem and goal of the project, as well as having
obtained the Olist eCommerce data, focus may now be shifted towards the
analyzing, cleaning, and transforming of the data. The first of which
will be the analysis of what all is included in the given data, and to
see what may be of use in further modelling down the line.

The analysis shall be done by adopting a two-stage analytical approach
to transition Olist from reactive to proactive logistics management.
Attempting to predict and in turn prevent delays before they occur, This
shall be done as such:

1.  **Unsupervised Learning (Strategic Seller Segmentation):** First,
    the features will be engineered to capture historical logistics
    performance and pricing behavior at the *seller level* (e.g.,
    average delay rate, average freight cost, dispatch speed). Then the
    **K-Means Clustering** shall be used to segment the sellers into
    distinct performance tiers (e.g., "High-Reliability," "Slow &
    Cheap," "High-Risk").
2.  **Supervised Learning (Tactical Order Forecasting):** Second, these
    sellers segments will be mapped back to the individual order level.
    These segments will be used as a key predictor in a classification
    model (such as a Random Forest or decision tree) to forecast whether
    a *future* individual order is likely to be delivered late.

**Business Impact** 

This pipeline enables a dual-pronged business strategy. Strategically,
Olist can use the seller clusters to penalize or coach "High-Risk"
sellers (addressing the root cause). Tactically, the predictive model
allows Olist's customer service team to identify high-risk orders the
moment they are placed, enabling proactive communication with the
customer or the deployment of expedited shipping interventions
(addressing the symptoms).

## Data dictionary

Having discussed various aspects of the data analysis process, the table
below may serve to illustrate a plethora of the variables which lie in
the database. With their ID's and descriptions outlined.

```{r data-table, echo=FALSE}
# Create a data dictionary of the merged dataset
data_dictionary <- data.frame(
  variable = names(orders_enriched),
  data_role = c(
    "identifier", "identifier", "identifier", "identifier",
    "temporal",
    "predictor", "predictor", "predictor", "predictor",
    "predictor", "predictor", "predictor",
    "context", "context", "context",
    "identifier", "context",
    "temporal", "temporal", "temporal", "temporal", "temporal",
    "predictor", "predictor",
    "identifier", "context", "context", "context",
    "outcome", "outcome"
  ),
  description = c(
    "Unique identifier for each order",
    "Unique identifier for the seller",
    "Unique identifier for the product",
    "Sequence number of item within the order",
    "Seller's shipping deadline for the item",
    "Listed price of the product",
    "Freight/shipping cost for the item",
    "Weight of the product in grams",
    "Length of the product in centimetres",
    "Height of the product in centimetres",
    "Width of the product in centimetres",
    "Product category name (English)",
    "Seller's zip code prefix",
    "Seller's city",
    "Seller's state",
    "Unique identifier for the customer (order-level)",
    "Order status at time of data extraction",
    "Timestamp when the order was placed",
    "Timestamp when payment was approved",
    "Date the order was handed to the carrier",
    "Date the order was delivered to the customer",
    "Estimated delivery date shown to customer",
    "Total payment value for the order (aggregated)",
    "Maximum instalment plan length for the order",
    "Anonymised unique customer identifier",
    "Customer's zip code prefix",
    "Customer's city",
    "Customer's state",
    "Target variable: whether the order was delivered late",
    "Signed delivery delay in days (negative = early)"
  )
)
knitr::kable(
  data_dictionary,
  format = "markdown",
  col.names = c("Variable", "Data Role", "Description")
)
```

# 3. Data Preparation

With the goal being to model an improved delays system in order to
increase customer satisfaction, and the variables now analysed, the
shift may be made towards preparing the data. From filtering unnecessary
or completely missing data, to restructuring, merging, or resorting
incomplete data, this chapter shall prepare the data for the Feature
Engineering.

## Dataset Overview and Selection

The Olist dataset comprises eight relational tables linked by shared
keys. We load all eight and make deliberate inclusion decisions based on
relevance to the business problem. The table below summarises these
decisions.

| Dataset | Included | Rationale |
|----------------------|------------------------|--------------------------|
| olist_orders | Yes | Analytical spine — contains all delivery timestamps and order status |
| olist_order_items | Yes | Links sellers to orders; provides price, freight, and product keys |
| olist_sellers | Yes | Seller geography — state and city for Stage 1 clustering |
| olist_customers | Yes | Customer state — used as a predictor in Stage 2 classification |
| olist_products | Yes | Physical dimensions (weight, size) — predictors of delivery complexity |
| olist_order_payments | Yes | Payment value and instalment behaviour — order-level predictors |
| olist_geolocation | No | Ambiguous deduplication required (multiple coordinates per zip code prefix); flagged as a direction for future improvement |
| olist_order_reviews | No | Outcome of delivery performance, not a predictor — inclusion would constitute data leakage. Used only for the proof-of-concept visualisation above. |

## Merge Strategy

The full merge sequence is:

1.  `orders` + `payments_aggregated` → via `order_id`
2.  
    -   `customers` → via `customer_id`
3.  `order_items` + `products_translated` → via `product_id`
4.  
    -   `sellers` → via `seller_id`
5.  Final join: orders spine + enriched order items → via `order_id`

## Column Removal

## Target Variable Construction

```{r target-variable-summary, echo=FALSE}
cat("On time:", sum(orders_enriched$is_late == "On_Time"), "\n")
cat("Late:   ", sum(orders_enriched$is_late == "Late"), "\n")
cat("Late rate:", round(mean(orders_enriched$is_late == "Late") * 100, 1), "%\n")
cat("Total labelled orders:", nrow(orders_enriched), "\n")
```

# 4. Data Exploration Philippe

## Missing values

```{r missing-values}
# Summary of missing values
missing_summary <- data.frame(
  variable = names(orders_enriched),
  n_missing = sapply(orders_enriched, function(x) sum(is.na(x))),
  pct_missing = sapply(orders_enriched, function(x) round(mean(is.na(x
)) * 100, 2))
)
missing_summary <- missing_summary[order(-missing_summary$pct_missing), ]
print(missing_summary)
```

Since these correspond to products with no physical measurements
recorded in the source they will be imputed with the median value for
each dimension.

```{r impute-dimensions}
dim_cols <- c("product_weight_g", "product_length_cm",
              "product_height_cm", "product_width_cm")

for (col in dim_cols) {
  orders_enriched[[col]][is.na(orders_enriched[[col]])] <-
    median(orders_enriched[[col]], na.rm = TRUE)
}

cat("Remaining NAs after imputation:", sum(is.na(orders_enriched)), "\n")
```

## Distributions and Skewness

### Numerical Variables

```{r outliers-skewness-distributions, echo=FALSE}

# LOG TRANSFORMATION CHECK for skewed vars
# Visual before/after for price, payment_value_total, freight_value, and product_weight_g 
par(mfrow = c(2, 2))
hist(orders_enriched$price, breaks = 50, main = "Price (raw)", xlab = "")
hist(log1p(orders_enriched$price), breaks = 50, main = "Price (log)", xlab = "")
hist(orders_enriched$payment_value_total, breaks = 50, main = "Payment (raw)", xlab = "")
hist(log1p(orders_enriched$payment_value_total), breaks = 50, main = "Payment (log)", xlab = "")
hist(orders_enriched$freight_value, breaks = 50, main = "Freight (raw)", xlab = "")
hist(log1p(orders_enriched$freight_value), breaks = 50, main = "Freight (log)", xlab = "")
hist(orders_enriched$product_weight_g, breaks = 50, main = "Weight (raw)", xlab = "")
hist(log1p(orders_enriched$product_weight_g), breaks = 50, main = "Weight (log)", xlab = "")
par(mfrow = c(1, 1))

```

### Categorical variables

```{r categorical-distributions, fig.height=5}
par(mfrow = c(1, 3))

state_counts <- sort(table(orders_enriched$customer_state), decreasing = TRUE)
barplot(state_counts, las = 2, main = "Orders by Customer State",
        col = "steelblue", cex.names = 0.7, ylab = "Count")

seller_state_counts <- sort(table(orders_enriched$seller_state), decreasing = TRUE)
barplot(seller_state_counts, las = 2, main = "Sellers by State",
        col = "coral", cex.names = 0.7, ylab = "Count")

cat_counts <- sort(table(orders_enriched$product_category_name_english),
                   decreasing = TRUE)
barplot(head(cat_counts, 20), las = 2, main = "Top 20 Product Categories",
        col = "steelblue", cex.names = 0.55, ylab = "Count")

par(mfrow = c(1, 1))
```

Both customer and seller activity are heavily concentrated in São Paulo
(SP), which accounts for approximately 40% of orders and the plurality
of sellers. This geographic concentration is relevant for our clustering
stage — SP-based sellers may exhibit systematically different logistics
profiles to sellers in more remote states. Product categories show a
similarly long-tail distribution, with bed/bath/table and health &
beauty dominating by volume. We retain the raw category variable as a
predictor in the classification model without further consolidation, as
Random Forest can handle high-cardinality categoricals natively.

## Data transformation

## Numerical variable transformations

```{r numerical-transformations}
# check skewness before transformation
skewness(orders_enriched$price, na.rm = TRUE)
skewness(orders_enriched$payment_value_total, na.rm = TRUE)
skewness(orders_enriched$freight_value, na.rm = TRUE)
skewness(orders_enriched$product_weight_g, na.rm = TRUE)
# Log transformation for skewed variables
orders_enriched$price_log <- log1p(orders_enriched$price)
orders_enriched$payment_value_total_log <- log1p(orders_enriched$payment_value_total)
orders_enriched$freight_value_log <- log1p(orders_enriched$freight_value)
orders_enriched$product_weight_g_log <- log1p(orders_enriched$product_weight_g)
# Check skewness after transformation
skewness(orders_enriched$price_log, na.rm = TRUE)
skewness(orders_enriched$payment_value_total_log, na.rm = TRUE)
skewness(orders_enriched$freight_value_log, na.rm = TRUE)
skewness(orders_enriched$product_weight_g_log, na.rm = TRUE)
```

# 5. Feature Engineering Ian/Philippe

## Seller-Level Feature Engineering

The first stage of our analytical pipeline requires building a
performance profile for each seller. Rather than analysing individual
orders in isolation, we aggregate historical order data to the seller
level, capturing each seller's logistics behaviour across all their past
orders. These features will serve as the input to our K-Means clustering
model.

We engineer four features that capture distinct dimensions of seller
performance:

-   **Late rate:** the proportion of a seller's orders delivered after
    the estimated date — our primary measure of reliability
-   **Avg dispatch days:** the average number of days between order
    purchase and carrier handoff — captures how quickly a seller
    processes and ships orders
-   **Avg freight cost:** the average shipping cost charged across a
    seller's orders — a proxy for logistics complexity and geographic
    reach
-   **Order volume:** the total number of fulfilled orders —
    distinguishes established sellers from low-activity ones whose
    metrics may be unreliable

```{r seller-feature-engineering}
# Build a seller-level performance profile from order-level data

# binary late flag at order level (numeric for aggregation)
orders_enriched$is_late_numeric <- ifelse(
  orders_enriched$is_late == "Late", 1, 0
)

# (days between purchase and carrier handoff)
orders_enriched$dispatch_days <- as.numeric(difftime(
  orders_enriched$order_delivered_carrier_date,
  orders_enriched$order_purchase_timestamp,
  units = "days"
))

# Aggregate to seller level
seller_profiles <- data.frame(
  seller_id = unique(orders_enriched$seller_id)
)

# Late delivery rate per seller
seller_profiles$late_rate <- tapply(
  orders_enriched$is_late_numeric,
  orders_enriched$seller_id,
  mean, na.rm = TRUE
)[seller_profiles$seller_id]

# Average dispatch speed per seller
seller_profiles$avg_dispatch_days <- tapply(
  orders_enriched$dispatch_days,
  orders_enriched$seller_id,
  mean, na.rm = TRUE
)[seller_profiles$seller_id]

# Average freight cost per seller
seller_profiles$avg_freight_log <- tapply(
  orders_enriched$freight_value_log,
  orders_enriched$seller_id,
  mean, na.rm = TRUE
)[seller_profiles$seller_id]

# Order volume per seller
seller_profiles$order_volume <- tapply(
  orders_enriched$is_late_numeric,
  orders_enriched$seller_id,
  length
)[seller_profiles$seller_id]

# Remove sellers with fewer than 10 orders
# Sellers with very few orders produce statistically unreliable rate estimates
# (e.g. 1 late delivery out of 2 orders = 50% late rate, which is misleading)
cat("Sellers before volume filter:", nrow(seller_profiles), "\n")
seller_profiles <- seller_profiles[seller_profiles$order_volume >= 10, ]
cat("Sellers after volume filter:", nrow(seller_profiles), "\n")
cat("Orders covered by remaining sellers:", 
    sum(orders_enriched$seller_id %in% seller_profiles$seller_id), "\n")

# Check for missing values in seller profiles
cat("\nMissing values in seller profiles:\n")
print(colSums(is.na(seller_profiles)))

# Inspect the distributions
summary(seller_profiles[, c("late_rate", "avg_dispatch_days", 
                              "avg_freight_log", "order_volume")])
```

The distributions of these seller features reveal important patterns
before we proceed to clustering:

```{r seller-profile-distributions}
par(mfrow = c(2, 2))

hist(seller_profiles$late_rate, breaks = 30,
     main = "Late Delivery Rate", 
     xlab = "Proportion of Late Orders",
     col = "steelblue")

hist(seller_profiles$avg_dispatch_days, breaks = 30,
     main = "Average Dispatch Speed",
     xlab = "Days from Purchase to Carrier",
     col = "steelblue")

hist(seller_profiles$avg_freight_log, breaks = 30,
     main = "Average Freight Cost",
     xlab = "R$ Freight Value",
     col = "steelblue")

hist(seller_profiles$order_volume, breaks = 30,
     main = "Order Volume",
     xlab = "Number of Orders",
     col = "steelblue")

par(mfrow = c(1, 1))
```

## Scaling for Clustering 

K-Means is a distance-based algorithm and is therefore sensitive to the
scale of input features. A seller with an average freight cost of R\$50
would dominate distance calculations over a late rate expressed as a
proportion between 0 and 1, simply due to the difference in units rather
than any meaningful difference in importance. We therefore standardise
all four features to zero mean and unit variance before clustering.

```{r seller-profile-scaling}
# Remove NAs before scaling
seller_profiles_clean <- na.omit(seller_profiles)
cat("Sellers lost to NA removal:", 
    nrow(seller_profiles) - nrow(seller_profiles_clean), "\n")

# Scale features (exclude seller_id)
seller_scaled <- scale(
  seller_profiles_clean[, c("late_rate", "avg_dispatch_days",
                              "avg_freight_log", "order_volume")]
)

# Confirm zero mean and unit variance
round(apply(seller_scaled, 2, mean), 6)
round(apply(seller_scaled, 2, sd), 3)
```

## PCA Feasibility Assessment

Before clustering, we assess whether dimensionality reduction via PCA is
appropriate. With only four features, PCA is unlikely to dramatically
simplify the input space — however it serves two purposes here: it
reveals how much variance each feature contributes, and it checks
whether any features are redundant (highly correlated), which would
justify exclusion before clustering.

```{r pca-assessment}
pca_result <- prcomp(seller_scaled, center = FALSE, scale. = FALSE)

# Variance explained by each component
pca_summary <- summary(pca_result)
print(pca_summary)

# Loadings: which features drive each component?
print(pca_result$rotation)

# Correlation matrix to check for redundancy
cor_matrix <- cor(seller_profiles_clean[, c("late_rate", "avg_dispatch_days",
                                              "avg_freight_log", "order_volume")])
print(round(cor_matrix, 3))
```

If the first two components explain more than 80% of variance, we use
the PCA scores as clustering input. If not — meaning all four features
contribute independently — we proceed with the full scaled feature set,
as discarding features would lose meaningful signal. We reflect on this
decision after inspecting the scree plot output.

## K-Means Seller Clustering

With seller profiles scaled and PCA assessed, we now cluster sellers
into performance tiers using K-Means. We determine the optimal number of
clusters using the elbow method on total within-cluster sum of squares.

```{r seller-clustering}
set.seed(42)

# Elbow method — test k from 1 to 10
wss <- sapply(1:10, function(k) {
  kmeans(seller_scaled, centers = k, nstart = 25)$tot.withinss
})

plot(1:10, wss, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Total Within-Cluster SS",
     main = "Elbow Method for Optimal k")

# Fit final model — adjust k based on elbow plot
set.seed(42)
k_optimal <- 4  # update after inspecting elbow plot
km_sellers <- kmeans(seller_scaled, centers = k_optimal, nstart = 25)

# Assign cluster labels back to seller profiles
seller_profiles_clean$cluster <- km_sellers$cluster

# Inspect cluster means to interpret what each cluster represents
cluster_means <- aggregate(
  cbind(late_rate, avg_dispatch_days, avg_freight_log, order_volume) ~ cluster,
  data = seller_profiles_clean,
  FUN = mean
)
print(round(cluster_means, 3))

# Cluster sizes
table(seller_profiles_clean$cluster)
```

```{r seller-cluster-interpretation}
# Visualise clusters using factoextra template
clusters <- km_sellers
cluster_plot_data <- seller_profiles_clean[, !names(seller_profiles_clean) %in% c("seller_id", "cluster"), drop = FALSE]

factoextra::fviz_cluster(
  object = clusters,
  data = cluster_plot_data,
  geom = "point",
  ellipse.type = "convex",
  ggtheme = theme_bw()
)

# Assign interpretable names based on cluster_means output

seller_profiles_clean$cluster_label <- dplyr::case_when(
  # bc of the high freight cost
  seller_profiles_clean$cluster == 1 ~ "Premium_Freight",
  # bc of the high late rate and slow dispatch
  seller_profiles_clean$cluster == 2 ~ "High_Risk",
  # bc of the low late rate and fast dispatch, even if freight is not the cheapest
  seller_profiles_clean$cluster == 3 ~ "High_Reliability",
  # bc of the high order volume but average late rate and dispatch speed
  seller_profiles_clean$cluster == 4 ~ "High_Volume",
  TRUE ~ "Unknown"
)

cat("\nSeller segment distribution:\n")
print(table(seller_profiles_clean$cluster_label))
```

## Mapping Seller Segments Back to Orders

The seller cluster labels are now joined back to the main orders
dataset, making the seller's performance tier available as a feature in
the classification model. This is the key link between Stage 1
(unsupervised seller segmentation) and Stage 2 (supervised order-level
prediction).

```{r map-clusters-to-orders}
# Join cluster labels onto orders
orders_enriched <- merge(
  orders_enriched,
  seller_profiles_clean[, c("seller_id", "cluster_label")],
  by = "seller_id",
  all.x = TRUE
)

# Orders from sellers excluded by the volume filter will have NA cluster
cat("Orders with cluster assigned:", 
    sum(!is.na(orders_enriched$cluster_label)), "\n")
cat("Orders without cluster (low-volume sellers):", 
    sum(is.na(orders_enriched$cluster_label)), "\n")

# Assign low-volume sellers their own segment rather than dropping them
orders_enriched$cluster_label[is.na(orders_enriched$cluster_label)] <- "Low_Volume"

# Check late rate by cluster — does segmentation separate risk levels?
aggregate(is_late_numeric ~ cluster_label, 
          data = orders_enriched, 
          FUN = mean)
```

The validation confirms the clustering has successfully separated seller
risk levels. The High_Risk segment shows a late delivery rate of 16.2% —
more than three times the 5.2% observed in the High_Reliability segment.
This suggests that the seller-level features we engineered are
meaningful predictors of delivery performance, and that our clustering
has effectively captured this signal in a way that can be leveraged for
order-level forecasting.

# 6. Modeling Giannis/Joeri

A

## Assumption Comparison

A

```{r modeling-data-split}
# Prepare modeling dataset with engineered features
model_data <- orders_enriched[, c(
  "is_late",              # target
  "cluster_label",        # seller segment — key engineered feature
  "price_log",            # transformed price
  "freight_value_log",    # transformed freight
  "payment_value_total_log", # transformed payment
  "payment_installments_max",
  "product_weight_g_log",
  "product_length_cm",
  "product_height_cm",
  "product_width_cm",
  "customer_state",
  "seller_state"
)]

model_data <- na.omit(model_data)
model_data$is_late <- factor(model_data$is_late, levels = c("On_Time", "Late"))

# Train/test split — 70/30, stratified on is_late
set.seed(42)
split <- initial_split(model_data, prop = 0.7, strata = "is_late")
train_data <- training(split)
test_data  <- testing(split)

cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:    ", nrow(test_data), "\n")
cat("Late rate (train):", round(mean(train_data$is_late == "Late") * 100, 1), "%\n")
cat("Late rate (test): ", round(mean(test_data$is_late == "Late") * 100, 1), "%\n")

# Convert categorical variables to factors for modeling
cat_cols <- c("cluster_label", "customer_state", "seller_state")
for (col in cat_cols) {
  train_data[[col]] <- as.factor(train_data[[col]])
  test_data[[col]]  <- as.factor(test_data[[col]])
}
```

## Model 1 — Decision Tree

A

```{r decision-tree-modeling}
# Fit a simple decision tree using C5.0
set.seed(42)
dt_model <- C50::C5.0(
  is_late ~ cluster_label + price_log + freight_value_log + payment_value_total_log
    + payment_installments_max + product_weight_g_log + product_length_cm
    + product_height_cm + product_width_cm + customer_state
    + seller_state, 
  data = train_data
)
print(dt_model)
# Predict on test set
dt_preds <- predict(dt_model, test_data)
# Confusion matrix
confusionMatrix(dt_preds, test_data$is_late, positive = "Late")
```

## Model 2 - Random Forest

A

```{r random-forest-modeling}
# Fit Random Forest with class weights to address imbalance
set.seed(42)
rf_model <- randomForest(
  is_late ~ .,
  data = train_data,
  ntree = 500,
  classwt = c("On_Time" = 1, "Late" = 10), # weight minority class
  importance = TRUE
)

print(rf_model)

# 5. Predict on test set
rf_preds <- predict(rf_model, test_data)
rf_probs <- predict(rf_model, test_data, type = "prob")[, "Late"]

# 6. Confusion matrix
confusionMatrix(rf_preds, test_data$is_late, positive = "Late")

# 7. Variable importance
varImpPlot(rf_model, main = "Random Forest — Variable Importance")

# aint looking good, so i'm testing out different threshold to see if the result can be better

# Test multiple thresholds and compare
thresholds <- c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30)

threshold_results <- data.frame(
  threshold   = thresholds,
  sensitivity = NA,
  specificity = NA,
  precision   = NA,
  balanced_acc = NA
)

for (i in seq_along(thresholds)) {
  preds <- as.factor(ifelse(rf_probs > thresholds[i], "Late", "On_Time"))
  preds <- factor(preds, levels = c("Late", "On_Time"))
  
  cm <- confusionMatrix(preds, test_data$is_late, positive = "Late")
  
  threshold_results$sensitivity[i]  <- cm$byClass["Sensitivity"]
  threshold_results$specificity[i]  <- cm$byClass["Specificity"]
  threshold_results$precision[i]    <- cm$byClass["Pos Pred Value"]
  threshold_results$balanced_acc[i] <- cm$byClass["Balanced Accuracy"]
}

print(round(threshold_results, 3))
# aight so this aint predicting shittt, but we can argue that if we set a low threshold (0.05) this can be used as a safety net to flag high-risk orders for manual review, even if it means a higher false positive rate. This is a common trade-off in imbalanced classification problems where the cost of missing a positive case (a late delivery) is much higher than the cost of a false alarm (flagging an on-time delivery as late). We do have to make the business case that the cost of false positives (e.g., unnecessary customer outreach, manual review time) is justified
```


## Model Reflection

A

## Model Output Comparison

A

# 7. Conclusion

A

# 8. Bibliography

Buck, A. (n.d.). Ecommerce share of retail sales [Updated 2026].
<https://www.mobiloud.com/blog/ecommerce-percentage-of-retail-sales>

Global ecommerce Sales Growth (2022–2028) [Updated Oct 2024]. (n.d.).
Oberlo.
<https://www.oberlo.com/statistics/global-ecommerce-sales-growth>
