---
title: "Group 21 Project"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

## Test loading the data
The loader code already has the capability to remove not useful columns, but we need to discuss which ones to remove. For now, we will load all the data and then decide which columns to keep or remove.
```{r run-main}
source("load.r")
# Load the data frames
# TODO remove not useful columns, it's already in the code but need to discuss which ones to remove
order_items_frame <- load_data("archive/olist_order_items_dataset.csv")
order_seller_frame <- load_data("archive/olist_sellers_dataset.csv")
order_product_frame <- load_data("archive/olist_products_dataset.csv")
order_translation_frame <- load_data("archive/product_category_name_translation.csv")
#display the data for testing
str(order_items_frame)
str(order_seller_frame)
str(order_product_frame)
```
## Create merged data frame
This merges the three data frames ( order_items_frame, order_seller_frame, order_product_frame) based on common columns (seller_id and product_id). The resulting merged data frame will contain all the relevant information from the three original data frames, allowing for more comprehensive analysis.
```{r merge-data}
# Merge the data frames based on common columns
merged_frame <- merge_frames(order_items_frame, order_seller_frame, by_column = "seller_id")
merged_frame <- merge_frames(merged_frame, order_product_frame, by_column = "product_id")

# Map Portuguese category names to English using a left join
merged_frame <- merge(
  merged_frame,
  order_translation_frame,
  by = "product_category_name",
  all.x = TRUE
)

# Keep only the English category name in the merged frame
merged_frame$product_category_name <- merged_frame$product_category_name_english
merged_frame$product_category_name_english <- NULL

# Display the structure of the merged data frame
str(merged_frame)
```

# Exploration
Now that we have the merged data frame, we can start exploring the data to gain insights and identify patterns. We can perform various analyses such as summary statistics, visualizations, and correlation analysis to understand the relationships between different variables in the dataset.
```{r exploration_boxplot}
# Summary statistics of the merged data frame
summary(merged_frame)

# boxplot of price by product category
boxplot(merged_frame$price ~ merged_frame$product_category_name,
        main = "Boxplot of Price by Product Category",
        xlab = "Product Category",
        ylab = "Price",
        las = 2, # Rotate x-axis labels for better readability
        cex.axis = 0.7) # Adjust x-axis label size
```

## Histogram of price to see the distribution of prices in the dataset
```{r exploration_histogram}
# Histogram of price
hist(merged_frame$price,
      main = "Histogram of Price",
      xlab = "Price",
      ylab = "Frequency",
      col = "lightblue",
      border = "black",
      breaks = 150)
```

## Looking at how often each product category appears in the dataset
```{r categorical_analysis}
cbind(table(merged_frame$product_category_name))

barplot(sort(table(merged_frame$product_category_name), decreasing = TRUE, las = 2),
        main = "Bar Plot of Product Categories",
        xlab = "Product Category",
        ylab = "Frequency",
        las = 2, # Rotate x-axis labels for better readability
        cex.axis = 0.7) # Adjust x-axis label size
```



