---
title: "Group 21 Project"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

## Test loading the data
The loader code already has the capability to remove not useful columns, but we need to discuss which ones to remove. For now, we will load all the data and then decide which columns to keep or remove.
```{r run-main}
source("load.r")
# Load the data frames
# TODO remove not useful columns, it's already in the code but need to discuss which ones to remove
order_items_frame <- load_data("archive/olist_order_items_dataset.csv")
order_seller_frame <- load_data("archive/olist_sellers_dataset.csv")
order_product_frame <- load_data("archive/olist_products_dataset.csv")
#display the data for testing
str(order_items_frame)
str(order_seller_frame)
str(order_product_frame)
```
## Create merged data frame
This merges the three data frames ( order_items_frame, order_seller_frame, order_product_frame) based on common columns (seller_id and product_id). The resulting merged data frame will contain all the relevant information from the three original data frames, allowing for more comprehensive analysis.
```{r merge-data}
# Merge the data frames based on common columns
merged_frame <- merge_frames(order_items_frame, order_seller_frame, by_column = "seller_id")
merged_frame <- merge_frames(merged_frame, order_product_frame, by_column = "product_id")

# Display the structure of the merged data frame
str(merged_frame)
```
