---
title: "Group 21 Project"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

## Test loading the data
The loader code already has the capability to remove not useful columns, but we need to discuss which ones to remove. For now, we will load all the data and then decide which columns to keep or remove.
```{r load-data}
source("load.r")
# Load the data frames
customer_frame <- load_data("archive/olist_customers_dataset.csv")

orders_frame <- load_data("archive/olist_orders_dataset.csv")
order_items_frame <- load_data("archive/olist_order_items_dataset.csv")
payment_frame <- load_data("archive/olist_order_payments_dataset.csv")

product_frame <- load_data("archive/olist_products_dataset.csv")
translation_frame <- load_data("archive/product_category_name_translation.csv")

```
## Create merged data frame
Here we will merge the data frames to create a single data frame that contains all the relevant information for our analysis. We will start by merging the product information with the order items, and then merge that with the orders and customer information. Finally, we will merge the payment information to have a complete view of each order. We will also remove any columns that are not relevant for our analysis to keep the data frame clean and focused on the variables we want to analyze.
```{r merge-data}
# Add English category names to products
products_translated <- merge(
  product_frame,
  translation_frame,
  by = "product_category_name",
  all.x = TRUE
)

# add columns for removal
columns_to_remove <- c("order_item_id", "shipping_limit_date", "seller_id")

# remove order_item_id since it only serves as an ID for the placement of the item in the order, which is not relevant for our analysis
order_items_frame <- remove_column(order_items_frame, columns_to_remove)

products_translated <- remove_column(products_translated, "product_category_name")

# Merge order items with translated product information 
order_items_enriched <- merge(
  order_items_frame,
  products_translated,
  by = "product_id",
  all.x = TRUE
)

columns_to_remove <- c("order_purchase_timestamp", "order_approved_at", "order_delivered_carrier_date", "order_delivered_customer_date", "order_estimated_delivery_date")
orders_frame <- remove_column(orders_frame, columns_to_remove)

# Merge orders with order items enriched with product information
orders_enriched <- merge(
  orders_frame,
  order_items_enriched,
  by = "order_id",
  all.x = TRUE
)

orders_enriched <- merge(
  orders_enriched,
  customer_frame,
  by = "customer_id",
  all.x = TRUE
)
# remove customer_id since customer_unique_id is more useful bc it allows us to identify unique customers, while customer_id is just an ID for the order, which is not relevant for our analysis
orders_enriched <- remove_column(orders_enriched, "customer_id")

# Aggregate payments to one row per order before merging bc there are multiple payments per order, which would create duplicate rows in the merged data frame and make analysis more difficult. By aggregating the payments to one row per order, we can have a cleaner and more accurate representation of the total payment value and installments for each order.
payments_by_order <- aggregate(
  cbind(payment_value_total = payment_value, 
        payment_installments_total = payment_installments) ~ order_id,
  data = payment_frame,
  FUN = sum
)

# Merge payments into the analytical table
orders_enriched <- merge(
  orders_enriched,
  payments_by_order,
  by = "order_id",
  all.x = TRUE
)
```

# Exploration
Now that we have the merged data frame, we can start exploring the data to gain insights and identify patterns. We can perform various analyses such as summary statistics, visualizations, and correlation analysis to understand the relationships between different variables in the dataset.
```{r exploration_boxplot}
# Summary statistics of the merged data frame
summary(orders_enriched)

# boxplot of price by product category
boxplot(orders_enriched$price ~ orders_enriched$product_category_name,
        main = "Boxplot of Price by Product Category",
        xlab = "Product Category",
        ylab = "Price",
        las = 2, # Rotate x-axis labels for better readability
        cex.axis = 0.7) # Adjust x-axis label size
```

## What are we trying to accomplish?
The goal is to identify economic demographics of customers based on area. In order to be able to predict emerging markets for certain product categories.

