---
title: "Group 21 Project"
output:
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup-data, include=FALSE}
source("helper.r")
source("merging.r")

# install and load packages
install_and_load()

# -- Load raw data -------------------------------------------------------
orders_raw      <- read.csv("archive/olist_orders_dataset.csv")
order_items_raw <- read.csv("archive/olist_order_items_dataset.csv")
payments_raw    <- read.csv("archive/olist_order_payments_dataset.csv")
customers_raw   <- read.csv("archive/olist_customers_dataset.csv")
products_raw    <- read.csv("archive/olist_products_dataset.csv")
sellers_raw     <- read.csv("archive/olist_sellers_dataset.csv")
translation_raw <- read.csv("archive/product_category_name_translation.csv")
reviews_raw     <- read.csv("archive/olist_order_reviews_dataset.csv")
geolocation_raw <- read.csv("archive/olist_geolocation_dataset.csv")

# merge and enrich orders dataset
orders_enriched <- create_enriched_orders(orders_raw, order_items_raw, payments_raw, 
                                          customers_raw, products_raw, sellers_raw, 
                                          translation_raw)

late_rate <- round(mean(orders_enriched$is_late == "Late") * 100, 1)
```

# 1. Introduction

## Business Context

The global e-commerce market continues to grow at a pace that outstrips
traditional retail. Brazil is a notable case: with a 15.4% year-on-year
growth rate, it ranked among the top five fastest-growing e-commerce
markets in 2023 (Oberlo, 2024). Within this rapidly expanding market,
Olist operates as a marketplace that connects small and medium-sized
sellers to customers across the country. Unlike vertically integrated
platforms, Olist does not own inventory or directly control fulfilment.
Instead, each seller manages their own warehousing, packaging, and
handoff to carriers independently. This decentralised model gives Olist
reach, but it also means that delivery performance is fundamentally
shaped by the operational behaviour of thousands of individual sellers.

When deliveries arrive late, it is the *customer* who experiences the
failure and the *platform* that absorbs the reputational damage.
Research on e-commerce satisfaction consistently identifies delivery
reliability as one of the strongest drivers of review scores and repeat
purchases. In the viewed dataset, this relationship is visible and
stark: late orders receive an average review score roughly two points
lower than on-time orders (the claim is validated in the
proof-of-concept visualisation below). This gap translates directly into
higher support costs, lower repeat purchase rates, and erosion of
platform trust.

## Problem Statement

Olist's current approach to late deliveries is purely reactive: customer
service intervenes only *after* the estimated delivery date has been
missed. This makes the process inherently suboptimal. Customers are
already dissatisfied by the time intervention begins, and the platform
has no systematic way to distinguish orders that are likely to arrive on
time from those that are at high risk of delay.

The root causes of late deliveries are not uniformly distributed across
the seller base. Some sellers consistently process and dispatch orders
quickly, while others exhibit chronic delays rooted in poor logistics
management, remote geography, or limited operational capacity. A purely
order-level approach to prediction would miss this structural dimension:
it would treat a late order from a consistently unreliable seller as
statistically equivalent to a one-off delay from an otherwise
high-performing seller. The understanding of *seller-level* patterns is
a prerequisite for effective *order-level* forecasting.

## Analytical Approach

A **two-stage analytical pipeline** is employed to separate the
structural (seller-level) and tactical (order-level) dimensions of the
problem.:

**Stage 1 - Unsupervised Seller Segmentation.** Features are engineered
to capture each seller's historical logistics performance (e.g., late
delivery rate, dispatch speed) and pricing behaviour (e.g., average
freight cost). These features are used in a K-Means clustering model to
segment sellers into distinct performance tiers. This stage addresses
the strategic question: *which sellers are structurally high-risk and
how their operational profiles differ?*

**Stage 2 - Supervised Order-Level Classification.** Seller segment
labels are mapped back to individual orders and combined with
order-level features such as product weight, price, payment method, and
geographic variables. These inputs serve as predictors in classification
models, including C5.0 decision trees and Random Forests, to forecast
the likelihood of late delivery. This stage addresses the tactical
question of *whether, at the time of purchase, an order can be
identified as high risk for delay.*

## Business Impact

This pipeline enables a dual-pronged strategy. **Strategically**, Olist
can use the seller clusters to identify chronically underperforming
sellers for targeted coaching, penalty structures, or enhanced
monitoring. This translates to addressing the root cause of delays.
**Tactically**, the order-level classification model allows customer
service to flag high-risk orders at the moment of purchase, enabling
proactive communication with the customer or the deployment of expedited
shipping, which means to address the symptoms before they materialise.

The intended audiences are therefore (1) Olist's **account management
team**, who would use the seller segmentation to inform seller
relationship strategies, and (2) Olist's **customer service
operations**, who would use the order-level predictions to prioritise
proactive outreach.

```{r proof-of-concept, echo=FALSE}


# Quick proof-of-concept: late delivery -> lower reviews
poc <- merge(orders_raw[, c("order_id",
                             "order_delivered_customer_date",
                             "order_estimated_delivery_date")],
             reviews_raw[, c("order_id", "review_score")],
             by = "order_id")
poc <- poc[!is.na(poc$order_delivered_customer_date) &
           !is.na(poc$order_estimated_delivery_date) &
           !is.na(poc$review_score), ]
poc$is_late <- ifelse(as.Date(poc$order_delivered_customer_date) >
                      as.Date(poc$order_estimated_delivery_date),
                      "Late", "On_Time")

review_impact <- aggregate(review_score ~ is_late, data = poc,
                           FUN = function(x) c(mean = round(mean(x), 2),
                                               n = length(x)))
review_impact <- do.call(data.frame, review_impact)
colnames(review_impact) <- c("Delivery_Status", "Avg_Review", "Count")

barplot(review_impact$Avg_Review,
        names.arg = review_impact$Delivery_Status,
        col = c("firebrick", "steelblue"),
        main = "Average Review Score: Late vs. On-Time Deliveries",
        ylab = "Average Star Rating (1-5)", ylim = c(0, 5))
text(x = c(0.7, 1.9), y = review_impact$Avg_Review + 0.2,
     labels = review_impact$Avg_Review)
```

# 2. Data Preparation

## Dataset Overview

The Olist dataset comprises nine relational tables linked by shared
keys. We load all nine and make deliberate inclusion decisions based on
relevance to the business problem.

Dataset -\> Included? -\> Rationale

`olist_orders` -\> Yes -\> Analytical spine - delivery timestamps, order
status `olist_order_items` -\> Yes -\> Links sellers to orders; provides
price, freight `olist_sellers` -\> Yes -\> Seller geography for Stage 1
clustering `olist_customers` -\> Yes -\> Customer state as a geographic
predictor `olist_products` -\> Yes -\> Physical dimensions - predictors
of delivery complexity `olist_order_payments` -\> Yes -\> Payment value
and instalment behaviour `product_category_name_translation` -\> Yes -\>
Maps Portuguese categories to English `olist_order_reviews` -\> Yes -\>
Outcome of delivery, not a predictor - inclusion would be data leakage.
Used only for the proof-of-concept above `olist_geolocation` -\> Yes -\>
Multiple coordinates per zip prefix require ambiguous deduplication;
flagged for future work

## Data dictionary

The table below may serve to illustrate a plethora of the variables
which lie in the database. With their ID's and descriptions outlined.

```{r data-dictionary, echo=FALSE}
dict <- data.frame(
  Variable = c("order_id", "seller_id", "product_id", "order_item_id",
               "shipping_limit_date", "price", "freight_value",
               "product_weight_g", "product_length_cm", "product_height_cm",
               "product_width_cm", "product_category_name_english",
               "seller_zip_code_prefix", "seller_city", "seller_state",
               "customer_id", "customer_unique_id",
               "customer_zip_code_prefix", "customer_city", "customer_state",
               "order_status", "order_purchase_timestamp", "order_approved_at",
               "order_delivered_carrier_date", "order_delivered_customer_date",
               "order_estimated_delivery_date",
               "payment_value_total", "payment_installments_max",
               "is_late", "delivery_delay_days"),
  Role = c("identifier", "identifier", "identifier", "identifier",
           "temporal", "predictor", "predictor",
           "predictor", "predictor", "predictor", "predictor",
           "context", "context", "context", "context",
           "identifier", "identifier",
           "context", "context", "predictor",
           "meta", "temporal", "temporal", "temporal", "temporal", "temporal",
           "predictor", "predictor",
           "outcome", "outcome"),
  Description = c(
    "Unique order identifier",
    "Unique seller identifier",
    "Unique product identifier",
    "Sequence number of item within order",
    "Seller's shipping deadline",
    "Listed product price (R$)",
    "Freight/shipping cost (R$)",
    "Product weight in grams",
    "Product length in cm",
    "Product height in cm",
    "Product width in cm",
    "Product category (English)",
    "Seller zip code prefix",
    "Seller city",
    "Seller state abbreviation",
    "Customer ID (order-level)",
    "Anonymised unique customer ID",
    "Customer zip code prefix",
    "Customer city",
    "Customer state abbreviation",
    "Order status at data extraction",
    "Timestamp: order placed",
    "Timestamp: payment approved",
    "Date: handed to carrier",
    "Date: delivered to customer",
    "Date: estimated delivery shown to customer",
    "Total payment value (aggregated across payment methods)",
    "Maximum instalment count for the order",
    "Target: Late or On_Time",
    "Signed delay in days (negative = early)")
)
knitr::kable(dict, format = "markdown")
```

## Merge

With the goal of modelling an improved delay management system to
increase customer satisfaction, and given the variables now analysed,
the focus shifts to data preparation. The primary objective in this
section is to merge data points that are relevant to each other within
the broader context of customer satisfaction, thereby enabling more
effective filtering in subsequent steps.

## Merge Strategy

The merge follows a deliberate sequence. Each table is first prepared
independently, after which they are joined in a controlled order to
preserve the correct analytical grain (one row per order-item
combination).

**Step 1: Aggregate payments to order level.** The payments table
contains multiple rows per order when a customer uses several payment
methods. These are aggregated to one row per order by summing the total
payment value and taking the maximum instalment count. This prevents row
inflation when joining to orders.

**Step 2: Translate product categories.** The translation table is
joined to the products table, replacing Portuguese category names with
their English equivalents. Products without a match are labelled
"Miscellaneous".

**Step 3: Enrich order items with product and seller information.**
Translated products and sellers are joined onto the `order_items` table
via `product_id` and `seller_id`, respectively.

**Step 4: Prepare the orders spine.** Date columns are parsed,
aggregated payments are joined via `order_id`, and customers are joined
via `customer_id`. The dataset is filtered to delivered orders only
(orders without delivery timestamps cannot be labelled), after which the
target variable is constructed.

**Step 5: Final join.** The enriched orders spine is joined to the
enriched `order_items` table via `order_id`, producing the
analysis-ready dataset.

```{r merge-pipeline, eval=FALSE}
# call merge function to show the full pipeline in one place
orders_enriched <- create_enriched_orders(orders_raw, order_items_raw, payments_raw, 
                                          customers_raw, products_raw, sellers_raw, 
                                          translation_raw)

cat("Final dataset dimensions:", nrow(orders_enriched), "rows,",
    ncol(orders_enriched), "columns\n")
```

## Target Variable Construction

```{r target-summary, echo=FALSE}
cat("On time:", sum(orders_enriched$is_late == "On_Time"), "\n")
cat("Late:   ", sum(orders_enriched$is_late == "Late"), "\n")
late_rate <- round(mean(orders_enriched$is_late == "Late") * 100, 1)
cat("Late rate:", late_rate, "%\n")
cat("Total labelled orders:", nrow(orders_enriched), "\n")
```

The late delivery rate of approximately `r late_rate`% means this is a
moderately imbalanced classification problem. The majority class
(on-time) outnumbers the minority class (late) by roughly 9:1. This has
direct implications for modelling: a naive classifier that always
predicts "On_Time" would achieve `r round(100 - late_rate, 1)`% accuracy
while being entirely ineffective for the business objective of
identifying at-risk orders. This imbalance is addressed explicitly in
the modelling stage through cost-sensitive learning and resampling.

# 3. Data Exploration

With an overview of the dataset established, the next step focuses on
exploring its internal structure. This analysis enables informed
decisions regarding data cleaning, filtering, and restructuring.
Specifically, variables are assessed for completeness, consistency, and
relevance, ensuring that only analytically meaningful data is retained.
This stage prepares the dataset for subsequent feature engineering and
modelling.

## Missing values

Missing values are handled through a structured approach. Orders with
partial missing information are retained where sufficient data remains
to describe the delivery process, particularly in cases where fulfillment
has occurred. This ensures that potentially valuable observations are
not discarded prematurely. In contrast, records with a substantial
proportion of missing variables are removed, as they do not provide
reliable analytically value. The number of such cases is minimal and does
not materially impact the data set.

```{r missing-values}
# Summary of missing values
missing_summary <- data.frame(
  variable = names(orders_enriched),
  n_missing = sapply(orders_enriched, function(x) sum(is.na(x))),
  pct_missing = sapply(orders_enriched, function(x) round(mean(is.na(x
)) * 100, 2))
)
missing_summary <- missing_summary[order(-missing_summary$pct_missing), ]
missing_summary <- missing_summary[missing_summary$n_missing > 0, ]
print(missing_summary)
```

The most common source of missing values is product physical dimensions
(`product_weight_g`, `product_length_cm`, `product_height_cm`,
`product_width_cm`). These correspond to a small number of products for
which physical attributes were not recorded. Because the missingness
rate is low and the variables are important predictors of delivery
complexity, median imputation is applied rather than dropping rows.
Median imputation is preferred over mean imputation because the
distributions are right-skewed (as shown below), making the median more
robust to outlier influence.

```{r impute-dimensions}
dim_cols <- c("product_weight_g", "product_length_cm",
              "product_height_cm", "product_width_cm")

for (col in dim_cols) {
  med_val <- median(orders_enriched[[col]], na.rm = TRUE)
  n_imputed <- sum(is.na(orders_enriched[[col]]))
  orders_enriched[[col]][is.na(orders_enriched[[col]])] <- med_val
  cat(col, ": imputed", n_imputed, "values with median =", med_val, "\n")
}
```

## Distributions and Skewness

With the missing variables eliminated a point of attention may be given
to the distribution and skewness of the data. As it would not be an
ideal scenario for individuals buying or selling extremely high or low
cost items to influence the value of products being delivered, or
alternatively for a singular individual who mass purchases a plethora of
items to skew the order amount.

### Distributions of Numerical Variables

Before deciding on transformations, the raw distributions of key
numerical predictors are inspected. Heavily right-skewed variables can
dominate distance-based calculations (K-Means) and influence split
selection in tree-based models, so identifying skewness at this stage
informs the log transformations applied in the next section.

```{r numerical-distributions, echo=FALSE, fig.height=8, fig.width=10}
par(mfrow = c(4, 2), mar = c(4, 4, 2, 1))

hist(orders_enriched$price, breaks = 50,
     main = "Price (raw)", xlab = "R$", col = "steelblue")
hist(log1p(orders_enriched$price), breaks = 50,
     main = "Price (log)", xlab = "log(1 + R$)", col = "steelblue")

hist(orders_enriched$freight_value, breaks = 50,
     main = "Freight (raw)", xlab = "R$", col = "coral")
hist(log1p(orders_enriched$freight_value), breaks = 50,
     main = "Freight (log)", xlab = "log(1 + R$)", col = "coral")

hist(orders_enriched$payment_value_total, breaks = 50,
     main = "Payment Total (raw)", xlab = "R$", col = "steelblue")
hist(log1p(orders_enriched$payment_value_total), breaks = 50,
     main = "Payment Total (log)", xlab = "log(1 + R$)", col = "steelblue")

hist(orders_enriched$product_weight_g, breaks = 50,
     main = "Product Weight (raw)", xlab = "grams", col = "coral")
hist(log1p(orders_enriched$product_weight_g), breaks = 50,
     main = "Product Weight (log)", xlab = "log(1 + g)", col = "coral")

par(mfrow = c(1, 1))
```

All four variables exhibit a skwewness, with a plethora of small data points spanning outwards driven by cheap or expensive orders, light or heavy products, and extensive shipping routes. In order to allow for a more evenly distributed dataset a log transformation is applied, this method has been chosen given data's skewness to the right. The log-transformed distributions (as seen in the right column) are far more symmetrical. The transformation `log1p()`, which mathematically is the natural logarithm of `(1 + x)`, is applied, as it handles zero values without issue and compresses the skewed data properly without losing rank-order information.In practical terms this removes the heavy influence which a small number of expensive, heave, and bulk orders have on the total dataset. 

```{r skewness-check}
cat("-- Skewness before transformation --\n")
cat("  price:              ", round(e1071::skewness(orders_enriched$price, na.rm = TRUE), 2), "\n")
cat("  freight_value:      ", round(e1071::skewness(orders_enriched$freight_value, na.rm = TRUE), 2), "\n")
cat("  payment_value_total:", round(e1071::skewness(orders_enriched$payment_value_total, na.rm = TRUE), 2), "\n")
cat("  product_weight_g:   ", round(e1071::skewness(orders_enriched$product_weight_g, na.rm = TRUE), 2), "\n")
```

### Distributions of Categorical Variables

With a more evenly distributed dataset now established, some preliminary insight may be given as to where the orders and sellers are distributed across Brazil. A simple visualization of this is made using the chunk below.

```{r categorical-distributions, fig.height=5}
par(mfrow = c(1, 3))

state_counts <- sort(table(orders_enriched$customer_state), decreasing = TRUE)
barplot(state_counts, las = 2, main = "Orders by Customer State",
        col = "steelblue", cex.names = 0.7, ylab = "Count")

seller_state_counts <- sort(table(orders_enriched$seller_state), decreasing = TRUE)
barplot(seller_state_counts, las = 2, main = "Sellers by State",
        col = "coral", cex.names = 0.7, ylab = "Count")

cat_counts <- sort(table(orders_enriched$product_category_name_english),
                   decreasing = TRUE)
barplot(head(cat_counts, 15), las = 2, main = "Top 15 Product Categories",
        col = "steelblue", cex.names = 0.55, ylab = "Count")

par(mfrow = c(1, 1))
```

What the chunk above shows is that buyer and seller activity are heavily concentrated in São Paulo (SP), accounting for approximately 40% of orders and of sellers. This geographic concentration may speak to larger socioeconomic topics, but for the scope of this report is left for the clustring stage. SP-based sellers may be prone to exhibiting different logistic strategies compared to sellers in more remote area's, where longer
distances to distribution hubs increase both freight costs and delivery
times. As for the product categories, these follow a distribution of
bed_bath_table, and health & beauty dominating by volume.

# 4. Data Transformation

## Numerical variable transformations

Based on the skewness observed above, `log1p()` is applied to the four
right-skewed numerical variables. This transformation is appropriate
because: (1) all values are non-negative, so `log1p()` is defined
everywhere; (2) the transformed distributions are substantially more
symmetric, improving the behaviour of distance-based algorithms
(K-Means) and reducing the dominance of extreme values in tree-based
model splits; and (3) the transformation is monotonic, preserving the
rank-order of observations.

```{r numerical-transformations}
# Log transformation for skewed variables
orders_enriched$price_log               <- log1p(orders_enriched$price)
orders_enriched$freight_value_log       <- log1p(orders_enriched$freight_value)
orders_enriched$payment_value_total_log <- log1p(orders_enriched$payment_value_total)
orders_enriched$product_weight_g_log    <- log1p(orders_enriched$product_weight_g)
# Check skewness after transformation
cat("-- Skewness after transformation --\n")
cat("  price_log:              ", round(e1071::skewness(orders_enriched$price_log, na.rm = TRUE), 2), "\n")
cat("  freight_value_log:      ", round(e1071::skewness(orders_enriched$freight_value_log, na.rm = TRUE), 2), "\n")
cat("  payment_value_total_log:", round(e1071::skewness(orders_enriched$payment_value_total_log, na.rm = TRUE), 2), "\n")
cat("  product_weight_g_log:   ", round(e1071::skewness(orders_enriched$product_weight_g_log, na.rm = TRUE), 2), "\n")
```

### Geographic Feature Engineering

The current feature set includes `customer_state` and `seller_state` as
categorical predictors, but these encode location independently, the
model knows *where* each party is, but not the *relationship* between
them. In a country as large as Brazil, the spatial distance between
seller and customer is likely a strong driver of delivery time: an order
shipped from São Paulo to a customer in the same city faces
fundamentally different logistics than one shipped from Rio Grande do
Sul to Amazonas.

Three geographic features are engineered to capture this relationship at
increasing levels of granularity:

1.  **`same_state`**: A binary indicator for whether the seller and
    customer are in the same state. Intra-state shipments typically
    benefit from shorter carrier routes and local distribution networks.

2.  **`cross_region`**: A binary indicator for whether the order crosses
    one of Brazil's five macro-regions (Norte, Nordeste, Centro-Oeste,
    Sudeste, Sul). Cross-region shipments often involve transfers
    between carrier networks and longer trunk routes, introducing
    additional delay risk.

3.  **`distance_km_log`**: The log-transformed Haversine distance (in
    km) between the seller's and customer's ZIP code centroids, computed
    from the `olist_geolocation` table. This provides a continuous
    measure of spatial separation. The geolocation table contains
    multiple coordinate records per ZIP prefix (representing different
    addresses within the same area); the median latitude and longitude
    per prefix are used to produce a single representative point,
    resolving the deduplication ambiguity flagged in Section 2.

```{r geographic_features}
# ============================================================
# Geographic Feature Engineering
# ============================================================

# --- Feature 1: Same-state indicator ---
orders_enriched$same_state <- ifelse(
  orders_enriched$customer_state == orders_enriched$seller_state, 1, 0
)

cat("Same-state orders:  ", sum(orders_enriched$same_state == 1), 
    "(", round(mean(orders_enriched$same_state) * 100, 1), "% )\n")
cat("Cross-state orders: ", sum(orders_enriched$same_state == 0), 
    "(", round(mean(orders_enriched$same_state == 0) * 100, 1), "% )\n")

# --- Feature 2: Cross-region indicator ---
# Brazil's five macro-regions
region_map <- c(
  AC = "Norte", AP = "Norte", AM = "Norte", PA = "Norte", 
  RO = "Norte", RR = "Norte", TO = "Norte",
  AL = "Nordeste", BA = "Nordeste", CE = "Nordeste", MA = "Nordeste", 
  PB = "Nordeste", PE = "Nordeste", PI = "Nordeste", RN = "Nordeste", 
  SE = "Nordeste",
  DF = "Centro-Oeste", GO = "Centro-Oeste", MT = "Centro-Oeste", 
  MS = "Centro-Oeste",
  ES = "Sudeste", MG = "Sudeste", RJ = "Sudeste", SP = "Sudeste",
  PR = "Sul", RS = "Sul", SC = "Sul"
)

orders_enriched$seller_region   <- region_map[orders_enriched$seller_state]
orders_enriched$customer_region <- region_map[orders_enriched$customer_state]
orders_enriched$cross_region    <- ifelse(
  orders_enriched$seller_region != orders_enriched$customer_region, 1, 0
)

cat("\nCross-region orders: ", sum(orders_enriched$cross_region == 1),
    "(", round(mean(orders_enriched$cross_region) * 100, 1), "% )\n")

# --- Feature 3: Haversine distance from geolocation table ---
# Deduplicate: one row per ZIP prefix (median coordinates)
geo_dedup <- aggregate(
  cbind(geolocation_lat, geolocation_lng) ~ geolocation_zip_code_prefix,
  data = geolocation_raw,
  FUN = median
)

# Join coordinates using match() instead of merge() to preserve all existing columns
seller_idx <- match(orders_enriched$seller_zip_code_prefix, 
                    geo_dedup$geolocation_zip_code_prefix)
orders_enriched$seller_lat <- geo_dedup$geolocation_lat[seller_idx]
orders_enriched$seller_lng <- geo_dedup$geolocation_lng[seller_idx]

customer_idx <- match(orders_enriched$customer_zip_code_prefix, 
                      geo_dedup$geolocation_zip_code_prefix)
orders_enriched$customer_lat <- geo_dedup$geolocation_lat[customer_idx]
orders_enriched$customer_lng <- geo_dedup$geolocation_lng[customer_idx]

# Haversine distance function
haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 6371  # Earth radius in km
  dlat <- (lat2 - lat1) * pi / 180
  dlon <- (lon2 - lon1) * pi / 180
  a <- sin(dlat / 2)^2 + 
       cos(lat1 * pi / 180) * cos(lat2 * pi / 180) * sin(dlon / 2)^2
  2 * R * asin(sqrt(a))
}

orders_enriched$distance_km <- haversine(
  orders_enriched$seller_lat, orders_enriched$seller_lng,
  orders_enriched$customer_lat, orders_enriched$customer_lng
)

# Log-transform (right-skewed, many short-distance orders)
orders_enriched$distance_km_log <- log1p(orders_enriched$distance_km)

# Report coverage and summary
cat("\nDistance computed for:", 
    sum(!is.na(orders_enriched$distance_km)), "orders\n")
cat("Missing (unmatched ZIP):", 
    sum(is.na(orders_enriched$distance_km)), "orders\n")
cat("\nDistance summary (km):\n")
print(summary(orders_enriched$distance_km))

# Region pair as a single interaction feature
orders_enriched$region_pair <- as.factor(paste(
  orders_enriched$seller_region, orders_enriched$customer_region, sep = "_"))
cat("\nRegion pairs:", nlevels(orders_enriched$region_pair), "unique combinations\n")
```

## Geographic Impact on Late Delivery

Before incorporating these features into the classification models,
their relationship with late delivery is examined directly. This
exploratory step serves two purposes: it validates that the features
carry predictive signal and it quantifies the magnitude of the
geographic effect, which informs the business recommendations.

```{r geographic_exploration}
# --- Late rate by same-state vs. cross-state ---
orders_enriched$is_late_numeric <- ifelse(
  orders_enriched$is_late == "Late", 1, 0
)
geo_impact <- aggregate(
  is_late_numeric ~ same_state,
  data = orders_enriched,
  FUN = function(x) c(late_rate = round(mean(x) * 100, 1), n = length(x))
)
geo_impact <- do.call(data.frame, geo_impact)
colnames(geo_impact) <- c("Same_State", "Late_Rate_Pct", "N_Orders")
geo_impact$Same_State <- ifelse(geo_impact$Same_State == 1, "Yes", "No")
cat("-- Late rate: Same-state vs. Cross-state --\n")
print(geo_impact)

# --- Late rate by cross-region ---
region_impact <- aggregate(
  is_late_numeric ~ cross_region,
  data = orders_enriched,
  FUN = function(x) c(late_rate = round(mean(x) * 100, 1), n = length(x))
)
region_impact <- do.call(data.frame, region_impact)
colnames(region_impact) <- c("Cross_Region", "Late_Rate_Pct", "N_Orders")
region_impact$Cross_Region <- ifelse(region_impact$Cross_Region == 1, "Yes", "No")
cat("\n-- Late rate: Same-region vs. Cross-region --\n")
print(region_impact)

# --- Late rate by region pair (seller -> customer) ---
# This shows which corridors are most problematic
orders_enriched$route <- paste(orders_enriched$seller_region, 
                               "->", 
                               orders_enriched$customer_region)

route_impact <- aggregate(
  is_late_numeric ~ route,
  data = orders_enriched,
  FUN = function(x) c(late_rate = round(mean(x) * 100, 1), n = length(x))
)
route_impact <- do.call(data.frame, route_impact)
colnames(route_impact) <- c("Route", "Late_Rate_Pct", "N_Orders")

# Only show routes with meaningful volume (>100 orders)
route_impact <- route_impact[route_impact$N_Orders >= 100, ]
route_impact <- route_impact[order(-route_impact$Late_Rate_Pct), ]

cat("\n-- Late rate by seller->customer region (>=100 orders) --\n")
print(route_impact, row.names = FALSE)

# --- Visualise: late rate vs. distance ---
# Bin distance into deciles for a clear visual
orders_with_dist <- orders_enriched[!is.na(orders_enriched$distance_km), ]
orders_with_dist$dist_bin <- cut(orders_with_dist$distance_km,
                                  breaks = quantile(orders_with_dist$distance_km, 
                                                     probs = seq(0, 1, 0.1)),
                                  include.lowest = TRUE, dig.lab = 4)

dist_late <- aggregate(
  is_late_numeric ~ dist_bin,
  data = orders_with_dist,
  FUN = mean
)

barplot(dist_late$is_late_numeric * 100,
        names.arg = seq(1, 10),
        xlab = "Distance Decile (1 = closest, 10 = farthest)",
        ylab = "Late Delivery Rate (%)",
        main = "Late Delivery Rate by Seller-Customer Distance",
        col = "steelblue")
```

The geographic analysis reveals a clear relationship between spatial
separation and late delivery risk. Cross-state orders exhibit a late
rate of 7.8%, nearly double the 4.4% observed for intra-state orders.
Cross-region orders show a similar pattern (7.5% vs. 6.0%), though the
effect is less pronounced because the region-level grouping is coarser.

The route-level analysis reveals that the effect is not purely a
function of distance but reflects regional logistics infrastructure.
Routes delivering *into* the Nordeste consistently exhibit the highest
late rates: Sudeste → Nordeste (12.8%), Sul → Nordeste (12.5%), and
Centro-Oeste → Nordeste (11.3%). Notably, even intra-Nordeste shipments
show elevated delay rates (9.0%), suggesting that carrier infrastructure
in the Northeast is structurally weaker, not merely disadvantaged by
distance. By contrast, routes within or into the Sul and Sudeste regions
achieve the lowest late rates (Sul → Sul: 3.8%, Centro-Oeste → Sudeste:
3.4%), reflecting the more developed logistics networks in southern
Brazil.

The distance decile plot confirms a broadly monotonic relationship
between seller-customer distance and late delivery probability, but the
route-level findings add an important nuance: distance alone is an
incomplete explanation. Two orders covering similar distances can face
very different delay risks depending on the *destination* region's
logistics capacity. This asymmetry has direct implications for Olist's
platform: delivery estimates and proactive interventions should account
for destination region, not just distance.

These findings have direct implications for Olist's platform:

-   **Customer-facing delivery estimates** could incorporate the
    seller-customer distance or region pair to set more accurate
    expectations at checkout, reducing the gap between promised and
    actual delivery dates that drives dissatisfaction.
-   **Seller recommendations**: When a customer searches for a product,
    the platform could prioritise sellers in the same region or state,
    where delivery reliability is structurally higher. This is not about
    hiding remote sellers but about transparently communicating delivery
    trade-offs.
-   **Carrier selection**: For high-risk corridors (e.g., cross-region
    routes with consistently elevated late rates), Olist could negotiate
    dedicated carrier partnerships or require expedited shipping
    options.

# 5. Feature Engineering and Dimensionality Reduction

## Seller-Level Feature Engineering

The first stage of the analytical pipeline requires building a
performance profile for each seller. Rather than analysing individual
orders in isolation, historical order data is aggregated to the seller
level, capturing each seller's logistics behaviour across past orders.
These seller-level features serve as input to the K-Means clustering
model.

Four features are engineered to capture distinct dimensions of seller
performance:

-   **Late rate:** the proportion of a seller's orders that arrived
    after the estimated delivery date - the primary reliability measure.
-   **Avg dispatch days:** the mean number of days between order
    purchase and carrier handoff - capturing operational processing
    speed.
-   **Avg freight cost (log):** the mean log-transformed freight cost -
    a proxy for logistics complexity and geographic reach.
-   **Order volume:** the total number of fulfilled orders -
    distinguishing established sellers from low-activity ones whose
    metrics may be statistically unreliable.

```{r seller-feature-engineering}
# Build a seller-level performance profile from order-level data

# Binary late indicator (numeric for aggregation)
orders_enriched$is_late_numeric <- ifelse(
  orders_enriched$is_late == "Late", 1, 0
)

# Dispatch speed: days from purchase to carrier handoff
orders_enriched$dispatch_days <- as.numeric(difftime(
  orders_enriched$order_delivered_carrier_date,
  orders_enriched$order_purchase_timestamp,
  units = "days"
))

# Aggregate to seller level using tapply
seller_profiles <- data.frame(
  seller_id = unique(orders_enriched$seller_id)
)

seller_profiles$late_rate <- tapply(
  orders_enriched$is_late_numeric, orders_enriched$seller_id,
  mean, na.rm = TRUE)[seller_profiles$seller_id]

seller_profiles$avg_dispatch_days <- tapply(
  orders_enriched$dispatch_days, orders_enriched$seller_id,
  mean, na.rm = TRUE)[seller_profiles$seller_id]

seller_profiles$avg_freight_log <- tapply(
  orders_enriched$freight_value_log, orders_enriched$seller_id,
  mean, na.rm = TRUE)[seller_profiles$seller_id]

# Order volume per seller
seller_profiles$order_volume <- tapply(
  orders_enriched$is_late_numeric,
  orders_enriched$seller_id,
  length
)[seller_profiles$seller_id]

# Remove sellers with fewer than 10 orders
# Sellers with very few orders produce statistically unreliable rate estimates
# (e.g. 1 late delivery out of 2 orders = 50% late rate, which is misleading)
cat("Sellers before volume filter:", nrow(seller_profiles), "\n")
seller_profiles <- seller_profiles[seller_profiles$order_volume >= 10, ]
cat("Sellers after volume filter:", nrow(seller_profiles), "\n")
cat("Orders covered by remaining sellers:", 
    sum(orders_enriched$seller_id %in% seller_profiles$seller_id), "\n")

# Check for missing values in seller profiles
cat("\nMissing values in seller profiles:\n")
print(colSums(is.na(seller_profiles)))

# Inspect the distributions
summary(seller_profiles[, c("late_rate", "avg_dispatch_days", 
                              "avg_freight_log", "order_volume")])
```

## Seller Profile Distributions

The distributions of these seller features reveal important patterns
before we proceed to clustering:

```{r seller-distributions, echo=FALSE, fig.height=6, fig.width=10}
par(mfrow = c(2, 2))

hist(seller_profiles$late_rate, breaks = 30,
     main = "Late Delivery Rate per Seller",
     xlab = "Proportion of Late Orders", col = "steelblue")

hist(seller_profiles$avg_dispatch_days, breaks = 30,
     main = "Average Dispatch Speed",
     xlab = "Days (Purchase -> Carrier)", col = "steelblue")

hist(seller_profiles$avg_freight_log, breaks = 30,
     main = "Average Freight Cost (log)",
     xlab = "log(1 + R$)", col = "steelblue")

hist(seller_profiles$order_volume, breaks = 30,
     main = "Order Volume per Seller",
     xlab = "Number of Orders", col = "steelblue")

par(mfrow = c(1, 1))
```

The distributions reveal meaningful variation across all four features.
The late rate distribution is right-skewed, with most sellers achieving
low late rates but a substantial tail of chronically poor performers.
Order volume is similarly skewed, with a few very high-volume sellers.
These distributions suggest that K-Means will identify clusters driven
primarily by the late rate and dispatch speed dimensions, which exhibit
the clearest separation between seller types.

## Scaling for Clustering

K-Means relies on Euclidean distance, which is sensitive to the scale of
input features. A seller with an average freight cost of R\$50 would
dominate distance calculations over a late rate expressed as a
proportion between 0 and 1, due to differences in units rather than
meaningful importance. All four features are therefore standardised to
zero mean and unit variance (Z-score standardisation) prior to
clustering.

```{r seller-profile-scaling}
# Remove NAs before scaling
seller_profiles_clean <- na.omit(seller_profiles)
cat("Sellers lost to NA removal:", 
    nrow(seller_profiles) - nrow(seller_profiles_clean), "\n")

# Scale features (exclude seller_id)
cluster_vars <- c("late_rate", "avg_dispatch_days",
                   "avg_freight_log", "order_volume")
seller_scaled <- scale(seller_profiles_clean[, cluster_vars])

# Verify standardisation
cat("\nMeans (should be ~0):\n")
print(round(colMeans(seller_scaled), 6))
cat("SDs (should be ~1):\n")
print(round(apply(seller_scaled, 2, sd), 3))
```

## PCA Feasibility Assessment

Before clustering, Principal Component Analysis is used to assess
whether the feature space can be simplified. With only four features,
substantial dimensionality reduction is unlikely; however, PCA serves
two purposes: (1) to reveal how variance is distributed across features,
and (2) to identify potential redundancy (high correlation), which could
justify feature exclusion prior to clustering.

```{r pca-assessment}
pca_result <- prcomp(seller_scaled, center = FALSE, scale. = FALSE)

# Variance explained by each component
pca_summary <- summary(pca_result)
print(pca_summary)

# Scree plot
plot(pca_result, type = "l", main = "Scree Plot - Seller Features")

# Correlation matrix to check for redundancy
cor_matrix <- cor(seller_profiles_clean[, cluster_vars])
cat("\nCorrelation matrix:\n")
print(round(cor_matrix, 3))

# Loadings, which features drive each component?
cat("\nPCA loadings:\n")
print(round(pca_result$rotation, 3))
```

The PCA results show that variance is distributed across all four
components, with no single component dominating overwhelmingly. The
correlation matrix confirms that the four features capture largely
independent dimensions of seller behaviour, as correlations are modest
and each feature contributes unique information. The full four-feature
set is therefore retained for clustering rather than reducing to PCA
scores, as discarding features would result in a loss of meaningful
signal without substantial gains in parsimony.

This decision aligns with the workbook guidance that dimensionality
reduction should be applied or deliberately avoided with sound
justification. In this case, reduction is *not* performed because (a)
the feature set is already small, (b) features are not redundant, and
(c) clustering on the original features preserves direct
interpretability of cluster profiles.

## K-Means Seller Clustering

Sellers are clustered into performance tiers using K-Means. The optimal
number of clusters is determined using the elbow method, which evaluates
how the within-cluster sum of squares decreases as the number of
clusters (k) increases.

```{r seller-clustering}
set.seed(42)

# Elbow method - test k from 1 to 10
wss <- sapply(1:10, function(k) {
  kmeans(seller_scaled, centers = k, nstart = 25)$tot.withinss
})

plot(1:10, wss, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Total Within-Cluster SS",
     main = "Elbow Method for Optimal k")
```

The elbow plot shows a clear bend at k = 4, after which additional
clusters provide diminishing reductions in within-cluster variance. Four
clusters are therefore selected, balancing granularity (sufficient
clusters to capture distinct seller types) with parsimony (few enough to
remain operationally meaningful and explainable to stakeholders).

```{r kmeans-fit}
set.seed(42)
k_optimal <- 4
km_sellers <- kmeans(seller_scaled, centers = k_optimal, nstart = 25)

seller_profiles_clean$cluster <- km_sellers$cluster

# Inspect cluster means in original (unscaled) space
cluster_means <- aggregate(
  cbind(late_rate, avg_dispatch_days, avg_freight_log, order_volume) ~ cluster,
  data = seller_profiles_clean,
  FUN = mean
)
cat("-- Cluster means (original scale) --\n")
print(round(cluster_means, 3))

cat("\n-- Cluster sizes --\n")
print(table(seller_profiles_clean$cluster))
```

## Cluster Visualisation and Interpretation

```{r cluster-visualisation, echo=FALSE, fig.cap="Two-dimensional PCA projection of seller clusters. Overlap in the projection is expected because the plot compresses four dimensions into two."}
factoextra::fviz_cluster(
  object = km_sellers,
  data   = seller_profiles_clean[, cluster_vars],
  geom   = "point",
  ellipse.type = "convex",
  ggtheme = ggplot2::theme_bw()
)
```

Based on the cluster means, interpretable labels are assigned. These
labels must be inspected and, if necessary, adjusted after examining the
actual `cluster_means` output, as K-Means cluster numbering is arbitrary
and may change across runs, even with the same seed, if the data
changes.

```{r cluster-labels}
# -- IMPORTANT: Inspect cluster_means output above and verify these --
# -- mappings match the actual cluster profiles before proceeding.   --
# The labels below are based on expected patterns; adjust as needed.

seller_profiles_clean$cluster_label <- dplyr::case_when(
  seller_profiles_clean$cluster == 1 ~ "Premium_Freight",
  seller_profiles_clean$cluster == 2 ~ "High_Risk",
  seller_profiles_clean$cluster == 3 ~ "High_Reliability",
  seller_profiles_clean$cluster == 4 ~ "High_Volume",
  TRUE ~ "Unknown"
)

cat("-- Seller segment distribution --\n")
print(table(seller_profiles_clean$cluster_label))
```

## Mapping Seller Segments Back to Orders

The seller cluster labels are joined back to the main orders dataset,
making each seller's performance tier available as a feature in the
Stage 2 classification models. This forms the key link between Stage 1
(unsupervised seller segmentation) and Stage 2 (supervised order-level
prediction).

```{r map-clusters-to-orders}
# Join cluster labels onto orders
orders_enriched <- merge(
  orders_enriched,
  seller_profiles_clean[, c("seller_id", "cluster_label")],
  by = "seller_id",
  all.x = TRUE
)

# Orders from sellers excluded by the volume filter will have NA cluster
cat("Orders with cluster assigned:      ",
    sum(!is.na(orders_enriched$cluster_label)), "\n")
cat("Orders without cluster (low-volume):",
    sum(is.na(orders_enriched$cluster_label)), "\n")

# Assign low-volume sellers their own segment rather than dropping them
orders_enriched$cluster_label[is.na(orders_enriched$cluster_label)] <- "Low_Volume"
orders_enriched$cluster_label <- as.factor(orders_enriched$cluster_label)
```

## Validating the Segmentation

A critical check before proceeding to classification is whether the
seller segmentation meaningfully separates late delivery risk. If the
late rate is similar across all clusters, the segmentation provides no
predictive value and the two-stage design is not justified.

```{r validate-segmentation}
# Check late rate by cluster - does segmentation separate risk levels?
seg_validation <- aggregate(
  is_late_numeric ~ cluster_label,
  data = orders_enriched,
  FUN = function(x) c(late_rate = round(mean(x) * 100, 1),
                       n_orders = length(x))
)
seg_validation <- do.call(data.frame, seg_validation)
colnames(seg_validation) <- c("Segment", "Late_Rate_Pct", "N_Orders")
print(seg_validation)
```

The validation confirms that the clustering successfully separates
seller risk levels. The High_Risk segment exhibits a substantially
higher late delivery rate than the High_Reliability segment. This
separation indicates that the engineered seller-level features capture
meaningful variation in delivery performance and that the cluster label
contributes predictive signal to the order-level classification models.

The ultimate test of whether the seller segment adds predictive value
beyond order-level features is conducted through an ablation test in the
modelling stage, where model performance is compared with and without
the `cluster_label` feature.

## Text Analytics: What Do Late-Delivery Reviews Reveal?

The proof-of-concept in Section 1 showed that late deliveries receive
substantially lower review scores; however, numerical ratings do not
reveal *what specifically* frustrates customers about delays. Topic
modelling is applied to compare themes in late-delivery versus on-time
reviews. This analysis serves two purposes: it validates the business
case with qualitative evidence and demonstrates how unstructured data
can enrich the seller management strategy.

### Preparing the Review Text

The Olist reviews are primarily in Portuguese. Since the lexicon-based
sentiment tools (`sentimentr`) are designed for English, a different
analytical approach is adopted: rather than computing sentiment scores,
delivery status (`is_late`) is used as a natural label that separates
negative context (late delivery) from positive context (on-time
delivery). Topic modelling is then applied to each group separately, and
the resulting themes are compared.

```{r text-merge-reviews}
# Merge reviews with delivery status
# Build one row per order label from the enriched table
order_labels <- unique(orders_enriched[, c("order_id", "is_late")])

review_text <- merge(
  reviews_raw[, c("order_id", "review_score", "review_comment_message")],
  order_labels,
  by = "order_id"
)

# Keep only reviews with actual text
review_text <- review_text[
  !is.na(review_text$review_comment_message) &
  review_text$review_comment_message != "", ]

cat("Reviews with text:", nrow(review_text), "\n")
cat("Late reviews:    ", sum(review_text$is_late == "Late"), "\n")
cat("On-time reviews: ", sum(review_text$is_late == "On_Time"), "\n")
```

```{r text-preprocess}
# Build a quanteda corpus
review_text$doc_id <- seq_len(nrow(review_text))
corpus_reviews <- quanteda::corpus(
  review_text,
  docid_field = "doc_id",
  text_field  = "review_comment_message"
)

# Tokenise: remove punctuation, numbers, symbols, URLs
tokens_reviews <- quanteda::tokens(
  corpus_reviews,
  what = "word",
  remove_punct   = TRUE,
  remove_symbols = TRUE,
  remove_numbers = TRUE,
  remove_url     = TRUE
)

# Lowercase
tokens_reviews <- quanteda::tokens_tolower(tokens_reviews)

# Remove Portuguese stopwords
tokens_reviews <- quanteda::tokens_remove(
  tokens_reviews,
  pattern = quanteda::stopwords("pt")
)

# Remove very short tokens (1-2 characters)
tokens_reviews <- quanteda::tokens_keep(
  tokens_reviews,
  min_nchar = 3
)

# Remove very short reviews (fewer than 5 tokens)
tokens_reviews <- tokens_reviews[quanteda::ntoken(tokens_reviews) >= 5]

# Build document-feature matrix
dfm_reviews <- quanteda::dfm(tokens_reviews)

# Trim rare terms (must appear in at least 15 documents)
dfm_reviews <- quanteda::dfm_trim(dfm_reviews, min_termfreq = 15)

# Remove empty documents after trimming
dfm_reviews <- quanteda::dfm_subset(dfm_reviews, quanteda::ntoken(dfm_reviews) > 0)

cat("Documents in DFM:", quanteda::ndoc(dfm_reviews), "\n")
cat("Features in DFM: ", quanteda::nfeat(dfm_reviews), "\n")
```

### Word Frequency Comparison

Before running topic models, the most frequent terms in late-delivery
and on-time reviews are compared, as this simple analysis often reveals
the dominant themes in each group.

```{r text-word-frequency, fig.width=12, fig.height=5}
# Split DFM by delivery status
dfm_late    <- quanteda::dfm_subset(dfm_reviews,
                quanteda::docvars(dfm_reviews, "is_late") == "Late")
dfm_ontime  <- quanteda::dfm_subset(dfm_reviews,
                quanteda::docvars(dfm_reviews, "is_late") == "On_Time")

# Top 15 terms per group
top_late   <- sort(colSums(dfm_late), decreasing = TRUE)[1:15]
top_ontime <- sort(colSums(dfm_ontime), decreasing = TRUE)[1:15]

par(mfrow = c(1, 2), mar = c(8, 4, 3, 1))
barplot(top_late, las = 2, col = "firebrick",
        main = "Top 15 Words - Late Deliveries",
        ylab = "Frequency", cex.names = 0.7)
barplot(top_ontime, las = 2, col = "steelblue",
        main = "Top 15 Words - On-Time Deliveries",
        ylab = "Frequency", cex.names = 0.7)
par(mfrow = c(1, 1))
```

The word frequency comparison provides an initial indication of what
differentiates the two groups. Late-delivery reviews are expected to
feature terms related to waiting, delays, and communication failures,
while on-time reviews are expected to emphasise product quality and
overall satisfaction.

### Topic Modelling: Late vs. On-Time Reviews

Separate LDA topic models are applied to late-delivery and on-time
reviews, revealing the themes that characterise each group by
identifying clusters of co-occurring terms rather than relying solely on
individual word frequencies.

```{r text-topic-models}
set.seed(42)

# Topic model on late-delivery reviews (k=2 topics)
lda_late <- seededlda::textmodel_lda(dfm_late, k = 2)

# Topic model on on-time reviews (k=2 topics)
lda_ontime <- seededlda::textmodel_lda(dfm_ontime, k = 2)

cat("-- Topics in LATE delivery reviews --\n")
print(seededlda::terms(lda_late, 10))

cat("\n-- Topics in ON-TIME delivery reviews --\n")
print(seededlda::terms(lda_ontime, 10))
```

```{r text-topic-sizes}
cat("-- Topic distribution (Late reviews) --\n")
print(table(seededlda::topics(lda_late)))

cat("\n-- Topic distribution (On-Time reviews) --\n")
print(table(seededlda::topics(lda_ontime)))
```

### Interpretation

The models reveal distinct thematic patterns that align with the
delivery experience captured by the `is_late` label.

**Late-delivery reviews** surface two interpretable topics. Topic 1
centres on frustration with the delivery process itself: terms such as
*ainda* ("still"), *dia* ("day"), *agora* ("now"), and *entregue*
("delivered") suggest customers actively tracking an overdue shipment
and expressing impatience about its status. Topic 2 shifts toward the
logistics chain and post-purchase evaluation, with *prazo* ("deadline"),
*correios* ("postal service"), and *loja* ("store") indicating that
customers attribute blame to both the carrier and the seller. The
presence of *recomendo* ("I recommend") in Topic 2 is notable, it likely
appears in negated form (*não recomendo*, "I do not recommend"),
reflecting the reputational damage that late deliveries inflict on
sellers. Together, these topics paint a picture of two distinct
frustration patterns: one focused on the *experience of waiting* and the
other on *assigning accountability* for the delay.

**On-time delivery reviews** tell a different story. Topic 1 is
product-focused, with *veio* ("it came"), *comprei* ("I bought"), and
*porém* ("however") suggesting customers evaluating the product itself,
including occasional caveats. Topic 2 is dominated by satisfaction
language: *antes* ("before" / "ahead of schedule"), *qualidade*
("quality"), *bom* ("good"), and *recomendo* ("I recommend") indicate
that timely delivery sets the stage for positive product evaluation and
endorsement. The term *antes* is particularly informative, it reveals
that customers notice and value deliveries that arrive *ahead* of the
estimated date, not merely on time.

The contrast between the two groups validates the business case in two
ways. First, it confirms that late deliveries trigger qualitatively
different review content, not simply lower scores but explicit
complaints about logistics failures. Second, it shows that on-time
delivery effectively "clears the stage" for product evaluation, meaning
that the delivery experience acts as a gatekeeper for overall customer
satisfaction. For Olist's account management team, this has a direct
implication: seller coaching should emphasise that reliable dispatch is
not just an operational metric but a prerequisite for positive customer
reviews.

A methodological limitation should be acknowledged. The reviews are in
Portuguese, and the topic model operates on raw Portuguese tokens
without lemmatisation or stemming. This means that morphological
variants (e.g., *entrega* / *entregue* / *entregar*) are treated as
separate features, potentially fragmenting topics. A more refined
pipeline would apply Portuguese-specific NLP preprocessing, which is
deferred to the limitations discussion in Section 7.

# 6. Modeling

Having established that seller segmentation separates delivery risk
levels (Section 5), the analysis proceeds to the supervised
classification stage. The objective is to predict, at the time of order
placement, whether an individual order will be delivered late. Beginning
with a single transparent model, inspecting its behaviour, exploring
strategies to modify that behaviour (boosting and cost-sensitive
learning), introducing an alternative model family (Random Forest),
evaluating all models under a shared cross-validation regime, and
combining them via hard voting.

## Modelling Assumptions and Approach Selection

The task is framed as a binary classification problem, where each order
is classified as either `Late` or `On_Time`. Two modelling families are
selected, each with distinct assumptions and trade-offs:

**C5.0 Decision Trees** are selected as the primary modelling family
because they produce transparent, rule-based classifications that can be
directly inspected and communicated to Olist's operations team. Decision
trees recursively partition the feature space using locally optimal
splits. Their key strengths in this context are interpretability (the
tree structure reveals *which* features drive late delivery risk) and
native handling of categorical predictors (seller segment, customer
state), while their main limitation is instability, as small changes in
the training data can produce substantially different trees. **C5.0
Decision Trees** are selected as the primary modelling family because
they produce transparent, rule-based classifications that can be
directly inspected and communicated to Olist's operations team. Decision
trees recursively partition the feature space using locally optimal
splits. Their key strengths in this context are interpretability (the
tree structure reveals *which* features drive late delivery risk) and
native handling of categorical predictors (seller segment, customer
state), while their main limitation is instability, as small changes in
the training data can produce substantially different trees.

**Random Forests** are selected as the second modelling family because
they address the instability limitation of single trees through
aggregation. A Random Forest builds multiple trees on bootstrap samples
with randomised feature subsets and combines their predictions via
majority vote, reducing sensitivity to individual data splits at the
cost of interpretability, a trade-off that is evaluated explicitly
below.

Deep learning is deliberately **not** implemented for this task, as the
dataset is tabular with well-defined features rather than
high-dimensional unstructured data (e.g., images or text). Neural
networks would introduce additional complexity without a clear
representational advantage, and the resulting loss of interpretability
would directly undermine the business objective of explaining risk
factors to Olist's account managers. This modelling choice is further
reflected upon in the Conclusion. Deep learning is deliberately **not**
implemented for this task, as the dataset is tabular with well-defined
features rather than high-dimensional unstructured data (e.g., images or
text). Neural networks would introduce additional complexity without a
clear representational advantage, and the resulting loss of
interpretability would directly undermine the business objective of
explaining risk factors to Olist's account managers. This modelling
choice is further reflected upon in the Conclusion.

## Preparing the Modelling Dataset

Features for the classification model are selected based on their
availability at the time of order placement (to avoid temporal leakage)
and their relevance to delivery performance. The `cluster_label` serves
as the key strategic feature, while the remaining predictors capture
order-level characteristics.

```{r modeling-data-split}
# Prepare modeling dataset with engineered features
model_data <- orders_enriched[, c(
  "is_late",                   # target
  "cluster_label",             # seller segment (Stage 1 output)
  "price_log",                 # order-level predictors
  "freight_value_log",
  "payment_value_total_log",
  "payment_installments_max",
  "product_weight_g_log",
  "product_length_cm",
  "product_height_cm",
  "product_width_cm",
  "customer_state",
  "seller_state",
  "same_state",                # NEW: geographic relationship features
  "cross_region",              # NEW
  "distance_km_log",            # NEW
  "region_pair"                # NEW: seller-customer region interaction
)]

model_data <- na.omit(model_data)

# Ensure correct types
model_data$is_late <- factor(model_data$is_late, levels = c("On_Time", "Late"))
model_data$cluster_label <- as.factor(model_data$cluster_label)
model_data$customer_state <- as.factor(model_data$customer_state)
model_data$seller_state   <- as.factor(model_data$seller_state)
model_data$same_state     <- as.factor(model_data$same_state)
model_data$cross_region   <- as.factor(model_data$cross_region)

cat("Modelling dataset:", nrow(model_data), "orders,",
    ncol(model_data), "variables\n")
cat("Late rate:", round(mean(model_data$is_late == "Late") * 100, 1), "%\n")
```

## Train-Test Split

The data is split 70/30 into training and test sets. Stratification on
the target variable is applied to preserve the class distribution in
both sets, ensuring that the minority class (Late) is adequately
represented in both training and evaluation.

```{r train-test-split}
# Train/test split - 70/30, stratified on is_late
set.seed(42)
split <- rsample::initial_split(model_data, prop = 0.7, strata = "is_late")
train_data <- rsample::training(split)
test_data  <- rsample::testing(split)

cat("Training set:", nrow(train_data), "orders\n")
cat("Test set:    ", nrow(test_data), "orders\n")
cat("Late rate (train):", round(mean(train_data$is_late == "Late") * 100, 1), "%\n")
cat("Late rate (test): ", round(mean(test_data$is_late == "Late") * 100, 1), "%\n")
```

## Addressing Class Imbalance

With approximately 90% of orders arriving on time, a naive classifier
that always predicts "On_Time" would achieve \~90% accuracy while
providing zero operational value. This class imbalance represents a
central challenge, which is addressed through two complementary
strategies:

1.  **Cost-sensitive learning** (applied to C5.0): higher
    misclassification costs are assigned to missed late deliveries
    (false negatives).
2.  **ROSE oversampling** (applied to the training set): a balanced
    training set is generated using the ROSE algorithm, which creates
    synthetic examples of the minority class; critically, this is
    applied only to the training data, while the test set retains the
    real-world distribution to ensure unbiased evaluation.

```{r class-imbalance}
# Show the imbalance problem explicitly
cat("-- Class distribution (training set) --\n")
print(table(train_data$is_late))
cat("Proportions:", round(prop.table(table(train_data$is_late)), 3), "\n")

# Apply ROSE oversampling to training data only
# NEVER apply to test data - test set must reflect real-world distribution
library(ROSE)
set.seed(42)
train_balanced <- ROSE::ROSE(is_late ~ ., data = train_data, seed = 42)$data

cat("\n-- Class distribution after ROSE --\n")
print(table(train_balanced$is_late))
cat("Proportions:", round(prop.table(table(train_balanced$is_late)), 3), "\n")
```

## Model 1 - Decision Tree

The modelling process begins with a single, unpruned C5.0 decision tree
trained on the ROSE-balanced data. This serves as a transparent
baseline, allowing the classification logic to be inspected rule by
rule. The C5.0 algorithm uses entropy-based splitting and produces
interpretable tree structures.

```{r decision-tree-modeling}
# Fit a simple decision tree using C5.0
set.seed(42)
dt_baseline <- C50::C5.0(
  is_late ~ .,
  data = train_balanced
)

# Print only the attribute usage (not the full tree)
cat("-- Tree size --\n")
cat("Nodes:", dt_baseline$size, "\n\n")
cat("-- Attribute usage --\n")
print(C50::C5imp(dt_baseline, metric = "usage"))

# Predict on ORIGINAL (unbalanced) test set
dt_base_preds <- predict(dt_baseline, test_data)
dt_base_probs <- predict(dt_baseline, test_data, type = "prob")[, "Late"]

cat("\n-- Confusion matrix: C5.0 Baseline --\n")
gmodels::CrossTable(
  test_data$is_late, dt_base_preds,
  prop.chisq = FALSE, prop.c = FALSE,
  prop.r = FALSE, prop.t = FALSE,
  dnn = c("Actual", "Predicted"))
```

The `summary()` output reveals the tree structure and attribute usage,
indicating which features the model relies on most. The `cluster_label`
and freight-related features are expected to appear prominently,
consistent with the domain understanding that seller behaviour and
logistics complexity are key drivers of delivery delay.

## Model 2 - Boosted C5.0 (10 Trials)

Boosting builds multiple trees sequentially, with each tree focusing on
cases misclassified by its predecessors. This approach can alter the
balance of classification errors. A total of 10 trials is used,
consistent with the workbook default.

```{r c50-boosted}
set.seed(42)
dt_boosted <- C50::C5.0(
  is_late ~ .,
  data = train_balanced,
  trials = 10
)

dt_boost_preds <- predict(dt_boosted, test_data)
dt_boost_probs <- predict(dt_boosted, test_data, type = "prob")[, "Late"]

cat("-- Confusion matrix: C5.0 Boosted (10 trials) --\n")
gmodels::CrossTable(
  test_data$is_late, dt_boost_preds,
  prop.chisq = FALSE, prop.c = FALSE,
  prop.r = FALSE, prop.t = FALSE,
  dnn = c("Actual", "Predicted"))
```

## Model 3 - Cost-Sensitive C5.0

In Olist's operational context, the two types of classification errors
have asymmetric consequences. A **false negative** (predicting On_Time
when the order is actually late) represents a missed opportunity for
proactive intervention, leading to unnotified customers, increased
support costs, and a high likelihood of negative reviews. A **false
positive** (predicting Late when the order arrives on time) results in
unnecessary proactive contact, which incurs operational effort but does
not cause reputational damage.

This asymmetry is formalised using a cost matrix, where a cost of 2 is
assigned to false negatives and a cost of 1 to false positives,
reflecting the judgment that missing a late delivery is twice as costly
as a false alarm.

```{r c50-cost-sensitive}
# Cost matrix: C5.0 expects rows = predicted, columns = actual
cost_matrix <- matrix(
  c(NA, 1,    # predicting Late  when actual On_Time -> FP cost = 1
    1.5,  NA),  # predicting On_Time when actual Late  -> FN cost = 2
  nrow = 2, ncol = 2, byrow = FALSE
)
rownames(cost_matrix) <- colnames(cost_matrix) <- c("On_Time", "Late")
cat("Cost matrix:\n")
print(cost_matrix)

set.seed(42)
dt_cost <- C50::C5.0(
  is_late ~ .,
  data = train_balanced,
  costs = cost_matrix
)

dt_cost_preds <- predict(dt_cost, test_data)
# Note: C5.0 does not support class probabilities when a cost matrix is
# used. We therefore cannot compute ROC/AUC for the cost-sensitive model.
# This is a known limitation of the C5.0 implementation, not a code error.
dt_cost_probs <- rep(NA_real_, nrow(test_data))

cat("\n-- Confusion matrix: C5.0 Cost-Sensitive --\n")
gmodels::CrossTable(
  test_data$is_late, dt_cost_preds,
  prop.chisq = FALSE, prop.c = FALSE,
  prop.r = FALSE, prop.t = FALSE,
  dnn = c("Actual", "Predicted"))
```

By assigning higher costs to false negatives, the model's behaviour
shifts toward more aggressive prediction of Late, capturing a greater
number of truly late orders at the expense of increased false positives.
The desirability of this trade-off depends on Olist's operational
capacity: if proactive customer contact is relatively inexpensive
compared to the reputational cost of missed late deliveries, the
cost-sensitive model is preferred; if contact capacity is constrained,
the baseline or boosted model may be more appropriate. This represents a
managerial decision rather than a purely statistical one.

## Model 4 - Random Forest

A Random Forest is introduced as a structurally different model family.
Random Forests build multiple trees on bootstrap samples with randomised
feature subsets and aggregate their predictions. The model is trained on
the ROSE-balanced data, consistent with the C5.0 models, and evaluated
on the original test set.

```{r random-forest-modeling}
set.seed(42)
rf_model <- ranger::ranger(
  is_late ~ .,
  data = train_balanced,  # balanced training data
  num.trees = 500,
  importance = "impurity",
  probability = TRUE
)
print(rf_model)

rf_probs <- predict(rf_model, test_data)$predictions[, "Late"]
rf_preds <- factor(
  ifelse(rf_probs > 0.5, "Late", "On_Time"),
  levels = c("On_Time", "Late")
)

cat("\n-- Confusion matrix: Random Forest --\n")
gmodels::CrossTable(
  test_data$is_late, rf_preds,
  prop.chisq = FALSE, prop.c = FALSE,
  prop.r = FALSE, prop.t = FALSE,
  dnn = c("Actual", "Predicted"))

# Variable importance
importance_df <- data.frame(
  variable = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
)

importance_df <- importance_df[order(-importance_df$importance), ]
barplot(importance_df$importance,
        names.arg = importance_df$variable,
        horiz = TRUE, las = 1,
        main = "Ranger - Variable Importance",
        xlab = "Mean Decrease Gini")
```

The variable importance plot reveals which features the Random Forest
relies on most for distinguishing late from on-time orders. This is
compared to the attribute usage reported by C5.0 to assess whether both
model families identify the same key drivers. Agreement between
structurally different models strengthens confidence in the identified
risk factors, while disagreement warrants closer inspection.

## Hyperparameter Tuning: Random Forest

The Random Forest above uses the `ranger` default for `mtry` (the number
of features randomly considered at each split), which is
$\lfloor\sqrt{p}\rfloor = 3$ for 15 predictors. To verify this default
is appropriate, a grid search over `mtry` values from 2 to 10 is
conducted, evaluating each configuration on the held-out test set.

```{r rf-mtry-tuning}
# mtry grid search results (hardcoded from tuning run to avoid re-fitting)
tuning_results <- data.frame(
  mtry     = c(2, 3, 4, 5, 6, 8, 10),
  AUC      = c(0.6867, 0.6860, 0.6843, 0.6832, 0.6825, 0.6804, 0.6796),
  Accuracy = c(0.7550, 0.7647, 0.7694, 0.7687, 0.7692, 0.7708, 0.7699),
  Recall   = c(0.4877, 0.4701, 0.4627, 0.4641, 0.4622, 0.4576, 0.4618),
  Precision= c(0.1314, 0.1332, 0.1344, 0.1343, 0.1342, 0.1341, 0.1345)
)
print(tuning_results)

barplot(tuning_results$AUC, names.arg = tuning_results$mtry,
        xlab = "mtry", ylab = "AUC", 
        main = "Random Forest AUC by mtry",
        col = ifelse(tuning_results$mtry == 2, "forestgreen", "grey70"),
        ylim = c(0.67, 0.69))
abline(h = tuning_results$AUC[tuning_results$mtry == 3], 
       lty = 2, col = "steelblue")
legend("topright", legend = "Default (mtry=3)", 
       lty = 2, col = "steelblue", cex = 0.8)
```

The grid search reveals that AUC varies by less than 0.008 across the
entire range of `mtry` values tested (0.6796 to 0.6867). The highest AUC
is achieved at `mtry = 2`, marginally outperforming the default
`mtry = 3` (0.6867 vs. 0.6860). This difference is small enough to be
within sampling variability, indicating that the Random Forest's
discriminative ability is robust to this hyperparameter. The default
value is therefore retained, as the marginal improvement does not
justify deviating from a well-understood baseline. This stability is
characteristic of Random Forests on datasets with moderate
dimensionality and is consistent with the workbook guidance that
ensemble methods are less sensitive to individual hyperparameter choices
than single models.

## ROC/AUC Comparison

So far, models have been evaluated using confusion matrices at a single
decision threshold, which obscures how well each model ranks orders by
their likelihood of being late. The ROC curve and AUC provide a
threshold-independent assessment of discriminative ability, indicating
whether the model assigns higher risk scores to orders that actually
arrive late.

```{r roc-comparison, fig.width=8, fig.height=6}
# Build ROC objects for models that produce probabilities
# (C5.0 with costs cannot output probabilities - excluded from ROC)
roc_base  <- pROC::roc(test_data$is_late, dt_base_probs,
                        levels = c("On_Time", "Late"), direction = "<")
roc_boost <- pROC::roc(test_data$is_late, dt_boost_probs,
                        levels = c("On_Time", "Late"), direction = "<")
roc_rf    <- pROC::roc(test_data$is_late, rf_probs,
                        levels = c("On_Time", "Late"), direction = "<")

# Plot ROC curves (3 models - cost-sensitive excluded)
plot(roc_base, col = "grey60", lwd = 2,
     main = "ROC Curves - Model Comparison")
plot(roc_boost, col = "steelblue", lwd = 2, add = TRUE)
plot(roc_rf,    col = "forestgreen", lwd = 2, add = TRUE)
legend("bottomright",
       legend = c(
         paste0("C5.0 Baseline (AUC = ", round(pROC::auc(roc_base), 3), ")"),
         paste0("C5.0 Boosted  (AUC = ", round(pROC::auc(roc_boost), 3), ")"),
         paste0("Random Forest (AUC = ", round(pROC::auc(roc_rf), 3), ")"),
         "C5.0 Cost - no probabilities (cost matrix)"
       ),
       col = c("grey60", "steelblue", "forestgreen", "firebrick"),
       lwd = c(2, 2, 2, 0), lty = c(1, 1, 1, 0),
       pch = c(NA, NA, NA, 4), cex = 0.8)
```

AUC values provide a threshold-independent measure of ranking quality. A
model with a higher AUC more effectively distinguishes late from on-time
orders *in general*, regardless of the chosen decision boundary.
However, AUC does not determine the optimal threshold; this choice
depends on Olist's operational capacity and cost tolerance, as discussed
above.

## Threshold Optimisation

All models above use the default decision threshold of 0.5: an order is
predicted as Late if the model assigns a probability above 50%. However,
with only 6.6% of orders arriving late, this threshold may not align
with Olist's operational priorities. Lowering the threshold increases
recall (more late orders are caught) at the expense of precision (more
false alarms), while raising it does the opposite.

The table below shows how the Random Forest's performance varies across
thresholds, illustrating the trade-off that Olist must navigate.

```{r threshold-analysis}
# Threshold analysis (hardcoded from tuning run)
thresh_results <- data.frame(
  Threshold = c(0.30, 0.35, 0.40, 0.45, 0.50, 0.55, 0.60),
  Accuracy  = c(0.4439, 0.5542, 0.6417, 0.7089, 0.7666, 0.8159, 0.8591),
  Recall    = c(0.8016, 0.7228, 0.6389, 0.5545, 0.4706, 0.3755, 0.2735),
  Precision = c(0.0882, 0.0998, 0.1112, 0.1219, 0.1344, 0.1468, 0.1613),
  F1        = c(0.1590, 0.1753, 0.1895, 0.1999, 0.2091, 0.2110, 0.2030)
)
print(thresh_results)

# Visualise the trade-off
par(mar = c(5, 4, 3, 4))
plot(thresh_results$Threshold, thresh_results$Recall, 
     type = "b", pch = 19, col = "firebrick", lwd = 2,
     xlab = "Decision Threshold", ylab = "Rate",
     main = "Precision-Recall Trade-off by Threshold",
     ylim = c(0, 0.85))
lines(thresh_results$Threshold, thresh_results$Precision, 
      type = "b", pch = 17, col = "steelblue", lwd = 2)
abline(v = 0.50, lty = 2, col = "grey50")
abline(v = 0.35, lty = 3, col = "firebrick")
legend("right", legend = c("Recall (Sensitivity)", "Precision", 
                            "Default (0.50)", "Proposed CS (0.35)"),
       col = c("firebrick", "steelblue", "grey50", "firebrick"),
       lwd = c(2, 2, 1, 1), lty = c(1, 1, 2, 3), pch = c(19, 17, NA, NA),
       cex = 0.8)
```

The Youden-optimal threshold (maximising sensitivity + specificity - 1)
is 0.412, slightly below the default. However, the optimal threshold
depends on the use case, not on a single statistical criterion.

For **Olist's customer service team**, a threshold of 0.35 is
recommended. At this threshold, the model catches 72% of truly late
orders (recall = 0.72), compared to only 47% at the default. The cost is
a lower precision (10.0% vs. 13.4%), meaning more false alarms -- but as
argued in the cost-sensitivity discussion, proactive contact with a
customer whose order will arrive on time is a minor operational cost,
while missing a late delivery generates genuine reputational damage.

For **Olist's account management team**, the default threshold of 0.50
or a slightly higher threshold of 0.55 is more appropriate. When the
consequence of a flag is a resource-intensive seller intervention,
precision matters more. At threshold 0.55, precision rises to 14.7%
while still identifying 38% of late orders.

## 5-Fold Cross-Validation

A single train-test split provides one estimate of model performance;
however, this estimate depends on the specific allocation of cases to
each set. To assess the **stability** of the results, all four models
are evaluated using 5-fold cross-validation with identical folds,
ensuring that observed differences reflect modelling choices rather than
evaluation artefacts.

```{r cross-validation, fig.width=10, fig.height=6}
set.seed(42)
k <- 5
folds <- caret::createFolds(model_data$is_late, k = k, list = TRUE)

# Storage: accuracy and AUC per fold per model
acc_mat <- matrix(NA, nrow = k, ncol = 4,
                  dimnames = list(NULL, c("C5.0_Base", "C5.0_Boost",
                                          "C5.0_Cost", "Random_Forest")))
auc_mat <- acc_mat  # same structure for AUC

# Storage for ensemble
all_preds <- vector("list", k)

for (i in seq_len(k)) {
  test_idx  <- folds[[i]]
  train_fold <- model_data[-test_idx, ]
  test_fold  <- model_data[test_idx, ]

  # Balance training fold with ROSE
  train_bal <- ROSE::ROSE(is_late ~ ., data = train_fold, seed = 42)$data

  # -- C5.0 Baseline --
  m1 <- C50::C5.0(is_late ~ ., data = train_bal)
  p1 <- predict(m1, test_fold)
  p1_prob <- predict(m1, test_fold, type = "prob")[, "Late"]
  acc_mat[i, "C5.0_Base"] <- mean(p1 == test_fold$is_late)
  auc_mat[i, "C5.0_Base"] <- as.numeric(pROC::auc(pROC::roc(
    test_fold$is_late, p1_prob, levels = c("On_Time", "Late"),
    direction = "<", quiet = TRUE)))

  # -- C5.0 Boosted --
  m2 <- C50::C5.0(is_late ~ ., data = train_bal, trials = 10)
  p2 <- predict(m2, test_fold)
  p2_prob <- predict(m2, test_fold, type = "prob")[, "Late"]
  acc_mat[i, "C5.0_Boost"] <- mean(p2 == test_fold$is_late)
  auc_mat[i, "C5.0_Boost"] <- as.numeric(pROC::auc(pROC::roc(
    test_fold$is_late, p2_prob, levels = c("On_Time", "Late"),
    direction = "<", quiet = TRUE)))

  # -- C5.0 Cost-Sensitive --
  m3 <- C50::C5.0(is_late ~ ., data = train_bal, costs = cost_matrix)
  p3 <- predict(m3, test_fold)
  acc_mat[i, "C5.0_Cost"] <- mean(p3 == test_fold$is_late)
  auc_mat[i, "C5.0_Cost"] <- NA_real_  # C5.0 with costs cannot produce probabilities

  # -- Random Forest --
  m4 <- ranger::ranger(is_late ~ ., data = train_bal,
                       num.trees = 500, probability = TRUE)
  p4_prob <- predict(m4, test_fold)$predictions[, "Late"]
  p4 <- factor(ifelse(p4_prob > 0.5, "Late", "On_Time"),
             levels = c("On_Time", "Late"))
  acc_mat[i, "Random_Forest"] <- mean(p4 == test_fold$is_late)
  auc_mat[i, "Random_Forest"] <- as.numeric(pROC::auc(pROC::roc(
    test_fold$is_late, p4_prob, levels = c("On_Time", "Late"),
    direction = "<", quiet = TRUE)))

  # Store predictions for ensemble
  all_preds[[i]] <- data.frame(
    actual  = test_fold$is_late,
    c50_base  = p1,
    c50_boost = p2,
    c50_cost  = p3,
    rf        = p4
  )
}

# -- Accuracy box plots --
par(mfrow = c(1, 2))
boxplot(acc_mat, main = paste0(k, "-Fold CV - Accuracy"),
        ylab = "Accuracy", col = c("grey80", "steelblue",
                                    "firebrick", "forestgreen"),
        las = 2, cex.axis = 0.8)

boxplot(auc_mat, main = paste0(k, "-Fold CV - AUC"),
        ylab = "AUC", col = c("grey80", "steelblue",
                               "firebrick", "forestgreen"),
        las = 2, cex.axis = 0.8)
par(mfrow = c(1, 1))

# Print summary statistics
cat("-- Mean Accuracy across folds --\n")
print(round(colMeans(acc_mat), 4))
cat("\n-- SD of Accuracy across folds --\n")
print(round(apply(acc_mat, 2, sd), 4))

cat("\n-- Mean AUC across folds --\n")
print(round(colMeans(auc_mat), 4))
cat("\n-- SD of AUC across folds --\n")
print(round(apply(auc_mat, 2, sd), 4))
```

The box plots reveal both the *level* and *stability* of each model's
performance. Models with similar average accuracy can differ
substantially in their variability across folds; a model with slightly
lower mean accuracy but a tighter spread may be preferable in practice
because it offers more predictable behaviour.

The AUC comparison is particularly informative because it is independent
of the decision threshold and directly measures how well each model
ranks orders by late-delivery risk. Differences in AUC that persist
across all five folds are more robust than those observed in a single
train-test split.

## Hard-Voting Ensemble

The four models are combined using hard voting (majority rule), where
each model casts a binary vote and the final prediction is determined by
the majority. In the case of a 2-2 tie, the prediction defaults to
"Late", consistent with the cost-sensitivity rationale that missing a
late delivery is more costly than a false alarm.

```{r ensemble, fig.width=10, fig.height=5}
# Hard voting function: 4 models, majority wins, ties -> "Late"
hard_vote <- function(pred_df) {
  late_votes <- rowSums(pred_df == "Late")
  factor(ifelse(late_votes >= 2, "Late", "On_Time"),   # tie -> Late
         levels = c("On_Time", "Late"))
}

# Evaluate ensemble across all CV folds
acc_ensemble <- numeric(k)

for (i in seq_len(k)) {
  fold_preds <- all_preds[[i]]
  pred_cols  <- fold_preds[, c("c50_base", "c50_boost", "c50_cost", "rf")]
  ensemble_pred <- hard_vote(pred_cols)
  acc_ensemble[i] <- mean(ensemble_pred == fold_preds$actual)
}

# Compare all models + ensemble
acc_all <- cbind(acc_mat, Ensemble = acc_ensemble)
boxplot(acc_all,
        main = paste0(k, "-Fold CV - Accuracy: Models vs. Ensemble"),
        ylab = "Accuracy",
        col = c("grey80", "steelblue", "firebrick",
                "forestgreen", "gold"),
        las = 2, cex.axis = 0.8)

cat("-- Ensemble mean accuracy --\n")
cat(round(mean(acc_ensemble), 4), "\n")
cat("-- Ensemble SD --\n")
cat(round(sd(acc_ensemble), 4), "\n")
```

Ensembling does not guarantee higher accuracy; its value lies in
*stability*. If the ensemble exhibits lower variability across folds
than individual models, it provides more predictable behaviour, which
may be operationally preferable even if peak accuracy is slightly lower.
Ensembling redistributes error rather than eliminating it, and whether
this redistribution is desirable depends on Olist's tolerance for
different types of mistakes.

## Ablation Test: Does Seller Segmentation Add Value?

The two-stage design rests on the claim that seller-level segmentation
provides predictive signal beyond what is captured by order-level
features alone. This claim is tested directly by re-running the Random
Forest (the strongest model) with and without the `cluster_label`
feature under the same cross-validation folds.

```{r ablation-test}
# Prepare data without cluster_label
model_data_no_cluster <- model_data[, !(names(model_data) %in% "cluster_label")]

# We already calculated these in the previous cross-validation chunk.
acc_with <- acc_mat[, "Random_Forest"]
auc_with <- auc_mat[, "Random_Forest"]

# Prepare storage for the "Without Cluster" runs
acc_without <- numeric(k)
auc_without <- numeric(k)

for (i in seq_len(k)) {
  test_idx   <- folds[[i]]

  # WITHOUT cluster_label
  train_wo <- ROSE::ROSE(is_late ~ .,
                          data = model_data_no_cluster[-test_idx, ],
                          seed = 42)$data
                          

  rf_wo <- ranger::ranger(is_late ~ ., data = train_wo,
                             num.trees = 500, probability = TRUE)
                             
  pwo_prob <- predict(rf_wo, model_data_no_cluster[test_idx, ])$predictions[, "Late"]
  pwo <- factor(ifelse(pwo_prob > 0.5, "Late", "On_Time"),
           levels = c("On_Time", "Late"))
  
  acc_without[i] <- mean(pwo == model_data_no_cluster[test_idx, "is_late"])
  auc_without[i] <- as.numeric(pROC::auc(pROC::roc(
    model_data_no_cluster[test_idx, "is_late"], pwo_prob,
    levels = c("On_Time", "Late"), direction = "<", quiet = TRUE)))
}

# Compare
ablation <- data.frame(
  Metric = c("Mean Accuracy", "SD Accuracy", "Mean AUC", "SD AUC"),
  With_Cluster    = c(round(mean(acc_with), 4), round(sd(acc_with), 4),
                      round(mean(auc_with), 4), round(sd(auc_with), 4)),
  Without_Cluster = c(round(mean(acc_without), 4), round(sd(acc_without), 4),
                      round(mean(auc_without), 4), round(sd(auc_without), 4))
)
print(ablation)

# Visual comparison
par(mfrow = c(1, 2))
boxplot(cbind(With = acc_with, Without = acc_without),
        main = "Ablation: Accuracy", ylab = "Accuracy",
        col = c("forestgreen", "grey70"))
boxplot(cbind(With = auc_with, Without = auc_without),
        main = "Ablation: AUC", ylab = "AUC",
        col = c("forestgreen", "grey70"))
par(mfrow = c(1, 1))
```

If the model with `cluster_label` consistently outperforms the model
without it across all five folds, this validates the two-stage design,
indicating that the unsupervised seller segmentation captures
information about delivery risk that is not redundant with order-level
features alone. If performance is similar, the segmentation may still
retain strategic value for seller management, even if it does not
improve tactical prediction.

## Model Comparison Summary

```{r model-summary}
# Compile results from the single train-test split
cat("==================================================\n")
cat("         MODEL COMPARISON (Test Set)              \n")
cat("==================================================\n\n")

models <- list(
  "C5.0 Baseline" = list(preds = dt_base_preds, probs = dt_base_probs),
  "C5.0 Boosted"  = list(preds = dt_boost_preds, probs = dt_boost_probs),
  "C5.0 Cost"     = list(preds = dt_cost_preds, probs = dt_cost_probs),
  "Random Forest"  = list(preds = rf_preds, probs = rf_probs)
)

for (name in names(models)) {
  preds <- models[[name]]$preds
  probs <- models[[name]]$probs

  cm <- caret::confusionMatrix(preds, test_data$is_late, positive = "Late")

  # AUC only if probabilities are available (not for cost-sensitive C5.0)
  if (all(is.na(probs))) {
    auc_val <- NA
  } else {
    auc_val <- as.numeric(pROC::auc(pROC::roc(
      test_data$is_late, probs,
      levels = c("On_Time", "Late"), direction = "<", quiet = TRUE)))
  }

  cat("-- ", name, " --\n")
  cat("  Accuracy:    ", round(cm$overall["Accuracy"], 4), "\n")
  cat("  Sensitivity: ", round(cm$byClass["Sensitivity"], 4),
      "  (recall for Late)\n")
  cat("  Specificity: ", round(cm$byClass["Specificity"], 4), "\n")
  cat("  Precision:   ", round(cm$byClass["Pos Pred Value"], 4), "\n")
  cat("  F1:          ", round(cm$byClass["F1"], 4), "\n")
  if (is.na(auc_val)) {
    cat("  AUC:          N/A (cost matrix - no probabilities)\n\n")
  } else {
    cat("  AUC:         ", round(auc_val, 4), "\n\n")
  }
}
```

## Model Reflection

The results reveal patterns that require careful interpretation rather
than a simple declaration of a "winning" model. The relevant question is
not *which model is right?* but *which risks are acceptable, and how
should disagreement be managed under uncertainty?*

**Accuracy alone is insufficient for imbalanced problems.** The Random
Forest achieves the highest accuracy (77.8%), but this metric
systematically rewards conservative behaviour. A model that rarely
predicts Late can achieve high accuracy by correctly classifying the
dominant On_Time class while providing limited operational value. A high
accuracy can coexist with operational blindness, particularly when the
event of interest is rare.

**The precision-recall trade-off reflects a decision problem, not a
modelling problem.** The four models span a clear spectrum: the
cost-sensitive C5.0 achieves the highest recall (73.1%), capturing
nearly three-quarters of truly late orders, but at a precision of only
8.6%, meaning most "Late" predictions are false positives. In contrast,
the Random Forest exhibits higher precision (13.7%) but lower recall
(44.6%), missing more than half of late deliveries. As highlighted in
Lecture 5, the appropriate position on this spectrum depends on the
relative costs of acting too frequently versus acting too late.

**Different stakeholders require different operating points.** This is
where the two-audience framing becomes decisive:

For **Olist's customer service team** (tactical, order-level), recall
should be prioritised. A false positive proactively contacting a
customer whose order will arrive on time, incurs at most a minor cost
and may even create a positive experience, while a false negative,
failing to identify a late delivery, leads to customer dissatisfaction,
low review scores, and reactive support costs. The *cost-sensitive C5.0*
model, with a recall of 73.1%, is therefore the recommended model for
this use case.

For **Olist's account management team** (strategic, seller-level),
precision is more important. Reaching out to a seller about performance
issues is a resource-intensive intervention that consumes account
manager time and may harm the relationship if the signal is incorrect.
In this context, the *Random Forest* or *boosted C5.0* models-being more
selective in their predictions-are better suited.

**Stability matters as much as performance level.** The cross-validation
results show that the Random Forest is the most stable model across
folds (accuracy SD = 0.005, AUC SD = 0.005), while the cost-sensitive
C5.0 exhibits the highest variability (accuracy SD = 0.017). For a
production system requiring predictable behaviour, the Random Forest
provides the most reliable baseline, whereas the cost-sensitive C5.0
offers higher recall at the cost of reduced stability across different
data samples.

**Ensembling redistributes error rather than eliminating it.** The
hard-voting ensemble moderates individual model predictions; ensembling
should be understood as a mechanism for managing how uncertainty is
expressed in decisions rather than as a guaranteed performance
improvement. It may reduce extreme failures, but does so by trading peak
performance for increased stability.

**The seller segment adds demonstrable value.** The ablation test
provides strong evidence for the two-stage design: including
`cluster_label` increases the Random Forest's mean AUC from 0.668 to
0.692 across all five folds, while also *reducing* variance (SD 0.005 vs
0.007). This indicates that seller segmentation captures delivery risk
information that is not redundant with order-level features alone. The
improvement is consistent rather than driven by a single split, thereby
validating the core analytical claim of the project.

# 7. Conclusion and Recommendations

## Summary of Findings

This project aims to shift Olist from reactive to proactive delivery
management through a two-stage analytical pipeline, and the results
validate this approach on both fronts.

**Stage 1 (Seller Segmentation)** identified four distinct seller
performance profiles using K-Means clustering on historical logistics
data. The segmentation effectively separates delivery risk, with the
High_Risk segment exhibiting a 16.2% late delivery rate-more than three
times the 5.2% observed among High_Reliability sellers. This separation
is not an artefact of the clustering algorithm but reflects genuine
differences in seller dispatch speed, freight cost structure, and
operational capacity.

**Stage 2 (Order-Level Classification)** demonstrated that late
deliveries can be predicted at the time of order placement using a
combination of seller segment membership, order characteristics, and
geographic features. The seller segment label (`cluster_label`) emerged
as the most important predictor, appearing in 100% of splits in the C5.0
baseline model. The ablation test confirmed this quantitatively, as
removing the seller segment reduced the Random Forest's mean AUC from
0.692 to 0.668 across all five cross-validation folds. The two-stage
design is therefore not merely a structural choice but contributes
measurable predictive value.

## Recommendations

The analytical pipeline yields recommendations at two levels, aligned
with the two-stage design and the two target audiences.

**For Olist's Account Management Team (Strategic - Seller-Level)**

The seller segmentation identifies a clearly separable High_Risk cluster
with a late delivery rate of 16.2%, more than three times the platform
average. Account managers should prioritise this segment for targeted
interventions, including structured performance reviews, logistics
coaching, and, where chronic underperformance persists, graduated
penalty structures such as reduced visibility in search rankings or
mandatory shipping insurance. The Premium_Freight cluster, despite low
late rates, exhibits elevated average freight costs; these sellers may
benefit from carrier renegotiation support or fulfilment hub
partnerships rather than performance penalties. Account managers should
also monitor Low_Volume sellers as they scale, since their segment
metrics are inherently less stable and may shift as order volume grows.

**For Olist's Customer Service Operations (Tactical - Order-Level)**

The classification models demonstrate that late delivery risk can be
estimated at the time of order placement with meaningful discriminative
ability (best AUC: 0.692). Customer service should adopt the
cost-sensitive C5.0 model for its operational alert pipeline, accepting
the higher false positive rate (specificity: 47.3%) in exchange for
capturing 71.9% of truly late orders. In practice, this means that for
flagged orders, customer service can send preemptive status updates,
offer delivery timeline extensions, or prepare expedited shipping
alternatives before the customer experiences frustration. For orders
that are not flagged, the standard reactive workflow remains
appropriate. The precise threshold for triggering proactive contact
should be calibrated in collaboration with Olist's operations team,
reflecting the available contact capacity and the marginal cost of each
outreach.

## Limitations

Several limitations qualify the conclusions drawn above.

First, **model performance remains modest in absolute terms**. The
best-performing model (Random Forest) achieves an AUC of 0.692 which is
substantially above random (0.5) but far from a reliable
individual-order prediction system. Precision for the Late class does
not exceed 13.4% in any model, meaning that the majority of flagged
orders will arrive on time. This limits the operational efficiency of
proactive contact and implies that the system is better suited as a
*prioritisation* tool (rank-ordering orders by risk) than as a binary
alerting system.

Second, **feature availability at prediction time is constrained**. The
model deliberately excludes post-purchase variables (e.g., carrier
handoff date, tracking events) to simulate prediction at the point of
sale. This design choice ensures operational relevance but removes
precisely the variables that carry the strongest predictive signal for
delivery outcomes. A production system with access to real-time
logistics data would likely outperform the models presented here.

Third, **the ROSE oversampling procedure introduces synthetic training
examples** that may not fully represent the structure of the minority
class. While ROSE was applied only to training data, preserving the
real-world distribution in the test set, the synthetic examples may
introduce noise that partly explains the high false positive rates
observed across all models.

Fourth, **the text analytics component operates on raw Portuguese
tokens** without lemmatisation, stemming, or translation. This limits
the granularity of the topic model and means that morphological variants
of the same concept are treated as distinct features. A more refined NLP
pipeline with Portuguese-specific preprocessing tools (e.g., the
`udpipe` package with a Portuguese model) would yield sharper topic
separation and more interpretable results.

Fifth, **temporal dynamics are not modelled**. The analysis treats the
dataset as a static cross-section, but in practice, seller performance
evolves over time. A seller classified as High_Risk in this analysis may
have improved (or deteriorated) during the observation period.
Incorporating temporal features, such as a recent trend in late rate or
time since first order, would better capture this evolution.

## Reflection on Advanced Techniques

This section addresses on how advanced techniques (deep learning, text
analytics, and reinforcement learning) could, in principle, be applied
to the problem. The focus is on what value they could add, what would be
required, and what constraints might limit their usefulness.

### Deep Learning

Deep learning, excels at learning rich representations from
high-dimensional unstructured data such as images. In this project,
however, the data is fundamentally tabular: structured features at the
order and seller level with well-defined meanings. Deep learning's
representational capacity offers limited advantage in this context for
several reasons.

First, the dataset contains approximately 110,000 observations with 11
predictors, a regime where tree-based models (C5.0, Random Forest) are
well-established as strong performers. Deep learning models typically
require substantially larger datasets to realise their representational
advantage and are prone to overfitting on smaller tabular datasets.
Second, interpretability is a core requirement of this project. The
business audiences (account managers and customer service) need to
understand *why* an order or seller is flagged, not just the prediction
itself. Decision trees and feature importance plots provide this
transparency directly; neural networks would require post-hoc
explanation methods (e.g., SHAP values) that add complexity without
clear benefit. Third, the feature space is already low-dimensional and
well-engineered, there are no raw, unprocessed inputs (such as images or
text sequences) that would benefit from the hierarchical feature
learning that deep learning provides.

One scenario where deep learning could add value is if the problem were
reformulated to incorporate unstructured data directly. For example, if
product images were available, a convolutional neural network could
learn visual features (e.g., product size, packaging quality) that
correlate with delivery complexity. Similarly, if raw review text were
modelled as a sequence (rather than as a bag of words), recurrent or
transformer-based architectures could capture more nuanced patterns.
However, both scenarios would require substantial additional data and
infrastructure that fall outside the scope of this project.

### Text Analytics

Text analytics was implemented in this project through LDA topic
modelling on Portuguese review text, comparing themes in late-delivery
versus on-time reviews. This implementation serves a complementary,
qualitative role: it validates the business case by showing that late
deliveries generate *qualitatively different* customer feedback, not
just lower scores.

The current implementation is deliberately basic. Reviews were tokenised
and filtered but not lemmatised, stemmed, or translated. More advanced
approaches could extend this analysis in several directions. Sentiment
analysis using Portuguese-specific lexicons or multilingual transformer
models (e.g., multilingual BERT) could quantify the intensity of
negative sentiment associated with late deliveries, moving beyond the
binary Late/On_Time grouping. Aspect-based sentiment analysis could
identify which specific aspects of the delivery experience (e.g.,
communication, packaging, carrier performance) drive dissatisfaction.
Additionally, review text features (e.g., sentiment scores, topic
proportions) could be integrated as predictors in the classification
models, potentially improving late-delivery prediction by capturing
customer-facing signals that structured order data does not.

However, incorporating text features into the prediction pipeline
introduces a conceptual complication: review text is generated *after*
delivery, meaning it cannot be used for prediction at the point of sale.
Text analytics therefore remains most valuable as a diagnostic and
strategic tool, helping Olist understand the *consequences* of late
delivery, rather than as a predictive input.

### Reinforcement Learning

Reinforcement learning (RL), addresses problems of sequential
decision-making under uncertainty, where an agent learns a policy by
interacting with an environment over time. The late-delivery prediction
problem, as currently framed, is a static classification task, each
order is evaluated independently at a single point in time and therefore
does not naturally lend itself to an RL formulation.

However, a meaningful RL application could emerge if the problem were
reformulated around Olist's *intervention decisions*. Consider the
following scenario: when an order is flagged as high-risk, Olist must
decide what action to take, send a preemptive notification, contact the
seller, offer expedited shipping, or do nothing. Each action has a cost,
and its effectiveness depends on the specific circumstances (seller
profile, product type, geographic distance). The outcome of the action
(whether the customer leaves a negative review, whether the delivery is
ultimately late) provides delayed feedback. This sequential decision
problem - repeatedly choosing interventions across thousands of orders
and learning which actions are most effective for which risk profiles,
maps naturally onto an RL framework.

In such a formulation, the *state* would include order features and the
seller's risk profile, the *actions* would be the available
interventions, and the *reward* would reflect a combination of customer
satisfaction (review scores), operational cost, and delivery outcome. A
Q-learning agent could learn which interventions are most cost-effective
for different types of orders, adapting its policy over time as the
platform accumulates more data. The key challenge would be defining a
realistic reward structure that balances competing objectives (e.g., the
cost of proactive contact versus the reputational cost of inaction),
that RL agents are highly sensitive to reward design choices.

This formulation would require real-time data on intervention outcomes,
which Olist does not currently make available in the public dataset and
would need careful consideration of exploration-exploitation trade-offs
in a setting where poor exploration choices (e.g., choosing not to
intervene on a truly late order) have real consequences for customers.
Nevertheless, it represents a conceptually sound extension that would
move Olist from *predicting* late deliveries to *optimally managing*
them.

## Final Remark

This project demonstrates that a two-stage analytical pipeline,
unsupervised seller segmentation followed by supervised order-level
classification, provides a coherent framework for shifting Olist's
delivery management from reactive to proactive. The seller segmentation
identifies structurally distinct performance profiles that are
operationally meaningful, and the ablation test confirms that this
segmentation adds measurable predictive value to the order-level models.
The text analytics component complements the quantitative pipeline by
revealing the qualitative consequences of late delivery in customers'
own words.

The models presented here are not production-ready, their precision is
too low for fully automated decision-making, and several data
limitations constrain their discriminative power. However, the
analytical design is sound: it separates the strategic question (*which
sellers are structurally high-risk?*) from the tactical question (*which
orders should be flagged?*), enabling different business units to act on
insights at the appropriate level. The path to deployment would involve
enriching the feature set with real-time logistics data, refining the
text analytics pipeline with Portuguese-specific NLP tools, and
calibrating decision thresholds in collaboration with Olist's operations
team.

# 8. Bibliography

Buck, A. (n.d.). Ecommerce share of retail sales [Updated 2026].
<https://www.mobiloud.com/blog/ecommerce-percentage-of-retail-sales>

Global ecommerce Sales Growth (2022-2028) [Updated Oct 2024]. (n.d.).
Oberlo.
<https://www.oberlo.com/statistics/global-ecommerce-sales-growth>
